---
title: "General Pharmaceutical Services England 2015/16 to 2021/22"
pagetitle: "GPhS England"
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

  <script>
  $(document).ready(function() {
    $('#TOC').prepend('<div><img src=`r knitr::image_uri("www/national-statistics-logo.svg")` alt=\"National Statistics logo\" style=\"position: relative; width: 50%; margin: 10% auto 10% auto; display: block;\"></div>');
  });

$(document).ready(function() {
  $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
});
</script>

  <style type="text/css">

  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }

* {
  font-family: Arial;
}

</style>
  <html lang="en">

```{r setup, include=FALSE}

# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
hcoptslang <- getOption("highcharter.lang")
  hcoptslang$thousandsSep <- ","
  hcoptslang$numericSymbols <- c("k","M","B","T","P","E")
  options(highcharter.lang = hcoptslang)


font <- "Arial"

colors <- c("#005EB8","#768692")

```

<main>

  <!-- # England 2015/16 to 2021/22 {.toc-ignore} -->

Published 13 October 2022




## Key findings

<!-- * There were 11,600 active community pharmacies and 112 active appliance contractors in England during 2021/22. 236 new pharmacies opened during 2021/22, while 451 closed. This is the lowest number of active contractors since 2015/16. -->

<!-- * 1.03 billion prescription items were dispensed by community pharmacies and appliance contractors in England in 2021/22. This is a 1.79% decrease from the number of items dispensed in 2021/22 but still a 2.35% increase in items dispensed since 2015/16. -->

<!-- * 964 million prescription items were dispensed via the Electronic Prescription Service (EPS) in 2021/22, 93.9% of all items dispensed in the year by community pharmacies and appliance contractors. This is an increase of 58.6 percentage points from 2015/16. -->

<!-- * The cost of drugs and appliances reimbursed to community pharmacies and appliance contractors totalled £8.97 billion in 2021/22. This was an increase of 3.72% from £8.65 billion in 2021/22 and a six year high despite the reduction in dispensed items in 2021/22. -->

<!-- * 2.77 million seasonal influenza vaccines were administered by community pharmacies in 2021/22. This was a 60.9% increase from the 1.72 million vaccines administered in  2021/2220 and a 365% increase on the 595 thousand vaccines administered in 2015/16. -->
<!-- * Covid vaccines -->



---

## 1. Introduction

  This publication is a National Statistic release. National Statistics status means that General Pharmaceutical Services (GPhS) - England meets the highest standards of trustworthiness, quality, and public value, and complies with all aspects of the [Code of Practice for Statistics.](https://code.statisticsauthority.gov.uk/) 
  
The designation of this publication as a National Statistic was confirmed in December 2021 following an assessment by the Office for Statistics Regulation (OSR). You can read more about this confirmation and the assessment of these statistics on the [OSR website.](https://osr.statisticsauthority.gov.uk/correspondence/ed-humpherson-to-matthew-wilson-confirmation-of-national-statistics-designation-for-statistics-on-general-pharmaceutical-services-in-england/)

### 1.1 Scope

From 1 April 2013, NHS England became responsible for the commissioning of NHS pharmaceutical services in England and for negotiating changes to arrangements for the provision of services. The [Community Pharmacy Contractual Framework](https://psnc.org.uk/contract-it/the-pharmacy-contract/) (CPCF) for pharmacy contractors is set out in the [The National Health Service (Pharmaceutical and Local Pharmaceutical Services) Regulations 2013](https://www.legislation.gov.uk/uksi/2013/349/contents/made), and the Pharmaceutical Services (Advanced and Enhanced Services) (England) Directions 2013 are contained within the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff).

In order to be able to provide pharmaceutical services for the NHS, the 2013 regulations state that a person (other than doctors or dentists) must be included in a pharmaceutical list. NHS England are required to prepare and maintain lists of those who have been granted applications. The list specifies both the premises and the named contractor.

To receive payment for the costs and fees incurred while providing services to the general public on behalf of the NHS, community pharmacies and appliance contractors must submit their prescriptions to NHSBSA along with a submission document, known as the FP34C. This has recently been digitised by the Manage Your Submission (MYS) application but was historically done via a paper form that accompanied paper prescription batches.

Data regarding advanced services such as the provision of medicines use reviews (MURs) and new medicines service (NMS) is taken from the FP34C submission document, or from the MYS application. This is then passed to NHSBSA payment systems to calculate account level payments that are to be made to community pharmacies and appliance contractors.

Data is collected from the submission of prescriptions by dispensing contractors to the NHSBSA. These prescriptions can be issued by GPs and other authorised prescribers such as nurses, dentists, and allied health professionals. Prescriptions that are issued by hospitals can also be dispensed in the community and submitted for reimbursement. Prescriptions that are issued in hospitals and fulfilled by the hospital pharmacy or dispensary are not included in this data.

Data is limited in this publication to only prescription items that have been dispensed by a community pharmacy or appliance contractor in England. Items dispensed by dispensing doctors, hospitals or prisons or submitted for reimbursement via a personal administration account have been excluded.

Data on advanced services provided by community pharmacies and appliance contractors is limited to where a claim has been submitted to NHSBSA in relation to performing one of those services. Counts around the number of services provided are based upon the number of times a fee has been paid to a contractor. The figures around the number of contractors that provide a service are based upon the number of contractors that have received payment of at least one fee for that service.


This publication aims  to describe the details of community pharmacy and appliance contractor activity in England across the whole financial year.

These statistics detail:

* the number of prescription items dispensed.
* the reimbursement costs of prescription items dispensed.
* the number of single activity fees they have received.
* the remuneration costs of single activity fees received.
* the number of essential and advanced services that they have provided.
* the associated costs for essential and advanced services.

These statistics cover the financial years between April 2015 and March 2022.


### 1.2 Definitions

A dispensing contractor or dispenser can be a community pharmacy or appliance contractor. Community pharmacies can dispense both drugs and appliances, but appliance contractors are limited to the supply of appliances as listed in Part IXA, IXB and IXC of the Drug Tariff.

Throughout this publication the term ‘item’, short for ‘prescription item’, refers to a single instance of a drug that is listed as a separate entry on a prescription form. For example, Paracetamol 500mg tablets x28 would be listed as one item, as would Ibuprofen 200mg tablets x56.

Prescription forms include both paper prescriptions and electronic messages sent via the Electronic Prescription Service (EPS). EPS allows prescribers to send prescriptions electronically to a dispenser (such as a pharmacy) of the patient's choice. This makes the prescribing and dispensing process more efficient and convenient for patients and staff.


There are many costs incurred when a dispensing contractor fulfils a prescription. The term cost refers to the basic price of the item and quantity prescribed that is reimbursed back to the contractor. This is sometimes called the 'Net Ingredient Cost' (NIC). Costs paid to contractors to cover the cost of dispensed items are referred to as reimbursement.

Fees are also payable to dispensing contractors for dispensing and for providing essential and advanced services to NHS patients. Many, but not all, fees are described in these summary statistics alongside their accompanying [summary tables](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx). Details of what fees are payable to community pharmacies and appliance contractors can be found in the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff). Fee payments to dispensing contractors for prescribing and other services are known as remuneration.

Further definitions of terms used in this publication can be found in the Glossary of the [Background Information and Methodology](https://www.nhsbsa.nhs.uk/sites/default/files/2021-10/gps_2021_background_information_methodology_v001.pdf) supporting document that accompanies this release.

---

## 2. Results and commentary

The charts in these statistics are interactive and allow you to isolate the topics that you are interested in by selecting or deselecting a series from the legend of the chart. For example, if you only want to view data for community pharmacies, you can click the other categories in the legend of the chart to deselect them. If you wish to include them again just click on their legend entry once more.

#### **`r paste0("Figure ",1,": Number of community pharmacies and appliance contractors between 2015/16 and 2021/22")`**

```{r total_contractors}
national_extract %>%
   dplyr::mutate(CONTRACTOR_TYPE =  case_when(APPLIANCE_DISPENSER_HIST == "Y" ~ "Appliance contractors",
                                  APPLIANCE_DISPENSER_HIST == "N" ~ "Community pharmacies")) %>% 
  dplyr::select(FINANCIAL_YEAR,num_contractors, CONTRACTOR_TYPE) %>% 
   group_by(FINANCIAL_YEAR, CONTRACTOR_TYPE) %>% 
  pivot_longer(cols = c(num_contractors),
               names_to = "measure",
               values_to = "value") %>% 
 
    group_chart_hc(
    x = FINANCIAL_YEAR,
    y = value,
    group = CONTRACTOR_TYPE,
    type = "line",
    xLab = "Financial year",
    yLab = "Number of contractors",
    title = ""
    
  )
```
Source: [Summary Tables - General Pharmaceutical Services - Table 1](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The number of community pharmacies and appliance contractors in England has remained fairly consistent for the past six years. However, the number of active community pharmacies in 2021/22 is the lowest it has been in the previous six years. This follows three consecutive years of decreases, with the number of community pharmacies decreasing by 3.76% between 2017/18 and 2021/22.

The number of appliance contractors in England changed in 2021/22 for the first time since 2018/19, decreasing by one.

#### **`r paste0("Figure ",2,": Total items dispensed by community pharmacies and appliance contractors between 2015/16 and 2021/22")`**

```{r items_dispensed}
national_extract %>%
   dplyr::mutate(CONTRACTOR_TYPE =  case_when(APPLIANCE_DISPENSER_HIST == "Y" ~ "Total items - appliance contractors",
                                  APPLIANCE_DISPENSER_HIST == "N" ~ "Total items - community pharmacies")) %>% 
  dplyr::select(FINANCIAL_YEAR,items_total, CONTRACTOR_TYPE) %>% 
   group_by(FINANCIAL_YEAR, CONTRACTOR_TYPE) %>% 
  pivot_longer(cols = c(items_total),
               names_to = "measure",
               values_to = "value") %>% 
 
    group_chart_hc(
    x = FINANCIAL_YEAR,
    y = value,
    group = CONTRACTOR_TYPE,
    type = "line",
    xLab = "Financial year",
    yLab = "Number of contractors",
    title = ""
    
  )
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The number of items dispensed by community pharmacies in England in between 2020/21 and 2021/22 increased by 2.59% from 1.02 billion to 1.04 billion. Overall, the number of items dispensed is still 4.80% higher than the 995 million items dispensed in 2015/16.

The number of items dispensed by appliance contractors in 2021/22 increased by 8.86% from 2020/21. This has increased from each year since 2015/16, from 7.91 million to 10.9 million items. This is an increase of 37.9%.

#### **`r paste0("Figure ",3,": Average number of items dispensed per contractor between 2015/16 and 2021/22")`**

```{r avg_items dispensed}
national_extract %>%
   dplyr::mutate(CONTRACTOR_TYPE =  case_when(APPLIANCE_DISPENSER_HIST == "Y" ~ "Appliance contractor",
                                  APPLIANCE_DISPENSER_HIST == "N" ~ "Community pharmacy"),
                 avg_items = items_total/num_contractors) %>% 
  dplyr::select(FINANCIAL_YEAR,avg_items, CONTRACTOR_TYPE) %>% 
   group_by(FINANCIAL_YEAR, avg_items) %>% 
  pivot_longer(cols = c(avg_items),
               names_to = "measure",
               values_to = "value") %>% 
 
    group_chart_hc(
    x = FINANCIAL_YEAR,
    y = value,
    group = CONTRACTOR_TYPE,
    type = "line",
    xLab = "Financial year",
    yLab = "Number of contractors",
    title = ""
    
  )

```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The yearly average number of items dispensed per contractor has been calculated by dividing the total number of items dispensed by the total number of contractors. This average is a mean.

The average number of items dispensed per contractor has increased by 3.60% from 87,400 in 2020/21 to 90,500 in 2021/22. This is due to a greater decrease in the total number of dispensed items compared to the decrease in dispensing contractors. There is still an overall increase of 8.68% from the average of 83,300 in 2015/16.

The average yearly items dispensed for appliance contractors has increased each year since 2015/16 and has now surpassed the average items dispensed for community pharmacies, at 98,300 in 2021/22. This is a 44.1% increase from 2015/16.

#### **`r paste0("Figure ",4,":Proportion of items dispensed via the Electronic Prescription Service (EPS) between 2015/16 and 2021/22")`**

```{r eps_items}

national_extract %>%
  dplyr::select(FINANCIAL_YEAR,items_total,eps_items_total_pharm) %>%
  dplyr::mutate(paper_items_all_cont = items_total - eps_items_total_pharm) %>%
  summarise(paper_items_all_cont = sum(paper_items_all_cont),
            eps_items_total_pharm = sum(eps_items_total_pharm),
            items_total = sum(items_total)) %>%
  tidyr::pivot_longer(cols = c(paper_items_all_cont,eps_items_total_pharm,items_total),
               names_to = "measure",
               values_to = "value") %>%
  dplyr::mutate(
    measure = case_when(measure == "paper_items_all_cont" ~ "Items dispensed by paper prescription",
                        measure == "eps_items_total_pharm" ~ "Items dispensed by EPS",
                        measure == "items_total" ~ "All items dispensed")
  ) %>%
  group_chart_hc(
    x = FINANCIAL_YEAR,
    y = value,
    group = "measure",
    type = "line",
    xLab = "Financial year",
    yLab = "Items dispensed",
    title = ""  )



```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

An NHS prescription can be fulfilled by either a paper form presented to a contractor by a patient, or by the Electronic Prescription Service (EPS). Since 2015/16 the proportion of prescription items that have been dispensed via EPS by community pharmacies and appliance contractors has increased significantly, and EPS is now the primary mechanism that prescriptions are issued and dispensed by.

In 2015/16, 355 million prescription items were dispensed via EPS, which was just 35.4% of all items in the year. In 2021/22, 1.00 billion prescription items were dispensed via EPS by community pharmacies and appliance contractors, which was 95.3% of all items in the year. This is an increase of 1.35 percentage points from 2020/21 and an increase of 59.9 percentage points between 2015/16 and 2021/22.

Additionally in 2021/22, 99.3% of all contractors dispensed at least one prescription item via EPS, an increase from 97.5% in 2015/16.

#### **`r paste0("Figure ",5,":Total costs and value of fees received for contractors between 2015/16 and 2021/22")`**
```{r costs_and_fees}
national_extract %>%
  dplyr::select(FINANCIAL_YEAR,costs_total,prof_fees_pharm) %>%
  summarise(costs_total = sum(costs_total),
            prof_fees_pharm = sum(prof_fees_pharm)) %>%
  tidyr::pivot_longer(cols = c(costs_total,prof_fees_pharm),
               names_to = "measure",
               values_to = "value") %>%
  dplyr::mutate(
    measure = case_when(measure == "costs_total" ~ "Total reimbursement",
                        measure == "prof_fees_pharm" ~ "Value of single activity fees")
  ) %>%
  group_chart_hc(
    x = FINANCIAL_YEAR,
    y = value,
    group = "measure",
    type = "line",
    xLab = "Financial year",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE  )
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

Contractors can receive many payments for fulfilling a prescription. These include being reimbursed for the basic price of the drug or appliance they have dispensed and receiving fees that relate to the dispensing of that item. There are many fees that can be claimed by a contractor depending on the item that is dispensed, details of what fees are payable to community pharmacies and appliance contractors can be found in the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff).

The 'single activity fee' (SAF) is a fixed fee that applies to every prescription item that is dispensed. The SAF was introduced in December 2016 by the Department of Health and Social Care (DHSC) to consolidate a range of payments into one single fee. These payments were:

* The professional fee (also known as dispensing fee)
* Practice payment
* Repeat dispensing payment
* EPS monthly allowance

The value of the SAF is set by DHSC and can fluctuate throughout the year. It is based upon the forecast of the total number of items dispensed to ensure that the fee delivery remains within the agreed funding envelope for the year. Table 1 displays the value of the SAF and the dates that it has changed.

#### **`r paste0("Table ",1,": Dispensing fees, their value, and when they came into effect")`**
``` {r saf_values}

saf_table <- data.frame(feeName = c("Professional fee","Single activity fee","Single activity fee","Single activity fee","Single activity fee","Single activity fee","Single activity fee"),
                        feeValue = c(90,113,125,129,126,127,129),
                        feeStart = c("01 Jan 2001","01 Dec 2016","01 Apr 2017","01 Nov 2017","01 Nov 2018","01 Aug 2019","01 Aug 2021"),
                        feeEnd = c("30 Nov 2016","31 Mar 2017","31 Oct 2017","31 Oct 2018","31 Jul 2019","31 Jul 2021","31 Mar 2022"))

DT::datatable(data = saf_table,
              rownames = FALSE,
              colnames = c("Fee name" = "feeName","Fee value (p)" = "feeValue","Fee start date" = "feeStart","Fee end date" = "feeEnd"),
              options = list(dom = "t",
                             columnDefs = list(list(orderable = FALSE,
                                               targets = '_all'))))

```

Source: [The Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff)

Figure 5 shows the total value of costs for drugs and appliances reimbursed to contractors, along with the value of all SAFs that have been claimed. This is not the total amount paid to contractors for dispensing an item as many fees are not included in these figures, such as additional fees that apply to controlled drugs. NHSBSA plans to include figures on all fees and payments that a contractor receives in future releases of this publication.

Costs reimbursed to contractors increased in 2021/22, for the third consecutive year. Costs increased by 0.83% between 2020/21 and 2021/22 from £8.97 billion to £9.05 billion. This is a seven year high.

The value of SAFs received by contractors increased in 2021/22 by 3.90% to £1.37 billion from £1.31 billion in 2020/21 which was a return to trend following a decrease in 2020/21 but following five years of increases 2015/16 to 2019/20. The value has increased from £920 million in 2015/16 to £1.31 billion in 2021/22, an increase of 48.4%. This has been driven by increases in the amount received for each SAF, from 90p in April 2015 to 129p in August 2021. Please note, prior to December 2016 the figures reported are for professional fees only, and do not include the three additional payments that the SAF consists of.

<!-- Another essential service contractors provide is the Discharge Medicine Service (DMS). This was introduced in February 2021 and allows patients discharged from NHS Trust hospitals who need extra support with their medicines to referred to their community pharmacy. -->

<!-- All open Pharmacy contractors on 1 February automatically received an upfront payment of £400 to support initial costs incurred in setting up the NHS Discharge Medicine Service (DMS) paid on 1 April 2021. -->

<!-- From February 2021 Pharmacies contractors will receive a fee of £35 for each completed NHS DMS On the rare occasion (see paragraph 7.4) that a pharmacy might not be able to complete the NHS DMS in full, they contractor will be able to claim the following fees for the stages of the NHS DMS that have been completed: -->

<!-- * Stage 1 - £12 -->

<!-- * Stage 2 - £11 -->

<!-- * Stage 3 - £12 -->

<!-- The total cost for Discharge Medicine Services in 2021/22 was £3.96 million.  -->

<!-- #### **`r paste0("Figure ",7,":Cost and fees for flu vaccinations provided by contractors between 2015/16 and 2021/22")`** -->
<!-- ```{r DISCHARGE_MEDS} -->

<!-- national_month_extract %>% -->
<!--   dplyr::select(YEAR_MONTH,DISCHARGE_MEDS) %>% -->
<!--   summarise(DISCHARGE_MEDS = sum(DISCHARGE_MEDS))  %>% -->
<!--   tidyr::pivot_longer(cols = DISCHARGE_MEDS, -->
<!--                       names_to = "measure", -->
<!--                       values_to = "values") %>% -->
<!--   dplyr::arrange(desc(measure)) %>% -->
<!--   mutate( -->
<!--     YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d") -->
<!--   ) %>% -->
<!--   dplyr::mutate( -->
<!--     measure = case_when(measure == "DISCHARGE_MEDS" ~ "Cost of vaccines") -->
<!--   ) %>% -->
<!--   group_chart_hc( -->
<!--     x = "YEAR_MONTH", -->
<!--     y = "values", -->
<!--     group = "measure", -->
<!--     type = "line", -->
<!--     xLab = "Month", -->
<!--     yLab = "Value (GBP)", -->
<!--     title = "", -->
<!--     currency = TRUE -->
<!--   ) %>% -->
<!--   hc_xAxis(type = "datetime") -->
<!-- ``` -->
### Advanced Services for Contractors

There are several Advanced Services within the NHS Community Pharmacy Contractual Framework (CPCF). Community pharmacies can choose to provide any of these services as long as they meet the requirements set out in the Secretary of State Directions.

#### Seasonal influenza vaccination advanced service

In 2015 community pharmacies began providing seasonal influenza vaccinations under a nationally commissioned service by NHS England & Improvement. Each year from September through to March pharmacy contractors can administer flu vaccines to patients and submit a claim to NHSBSA for payment. This includes reimbursement of the cost of the vaccine, plus a fee for providing the service to NHS patients.

The figures in these statistics for flu vaccinations only relate to those administered by community pharmacies, and do not include vaccines issued to GP patients, healthcare workers, or children of primary school age. [Official Statistics for these are produced by Public Health England](https://www.gov.uk/health-and-social-care/health-protection-immunisation).

#### **`r paste0("Figure ",6,":Number of flu vaccinations provided by contractors between 2015/16 and 2021/22")`**

```{r flu_items}

flu_data %>%
  dplyr::select(FINANCIAL_YEAR,flu_items_total_csv) %>%
  summarise(flu_items_total_csv = sum(flu_items_total_csv)) %>% 
  basic_chart_hc(
    x = "FINANCIAL_YEAR",
    y = "flu_items_total_csv",
    type = "line",
    xLab = "Financial year",
    yLab = "Flu vaccines",
    title = ""
  )

```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The number of vaccines administered by pharmacies as part of the Flu Vaccination Advanced Service has increased by 8.15 times since the service's inception in 2015/16. In 2021/22 there were  4.85 million vaccines  administered by 9,860 community pharmacies, an increase of 75.1% on 2020/20,  at an average of 492 vaccines per pharmacy.


#### **`r paste0("Table ",2,": Average number of flu vaccines provided per community pharmacy that provides the advanced service between 2015/16 and 2021/22")`**
``` {r avg_flu_items_table}
flu_table <- flu_data %>%
  dplyr::select(FINANCIAL_YEAR,flu_items_total_csv,num_flu_pharm_csv) %>%
  summarise(flu_items_total_csv = sum(flu_items_total_csv),
            num_flu_pharm_csv = sum(num_flu_pharm_csv)) %>% 
  dplyr::mutate(avg_flu_items = round(flu_items_total_csv/num_flu_pharm_csv,0)) %>%
  tidyr::pivot_longer(
    cols = c(flu_items_total_csv,num_flu_pharm_csv,avg_flu_items),
    names_to = "Measure",
    values_to = "values"
  ) %>%
  filter(Measure == "avg_flu_items") %>%
  dplyr::mutate(
    Measure = case_when(Measure == "avg_flu_items" ~ "Average number of flu vaccines per community pharmacy")
  ) %>%
  pivot_wider(
    names_from = FINANCIAL_YEAR,
    values_from = values
  )


DT::datatable(data = flu_table,
              rownames = FALSE,
              options = list(dom = "t",
                             columnDefs = list(list(orderable = FALSE,
                                                    targets = '_all'))))
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The average number of flu vaccines administered per community pharmacy has been calculated by dividing the total number of flu vaccines administered for each year by the total number of community pharmacies that have provided the service. This average is a mean.

The average number of flu vaccines administered per community pharmacy that provides the seasonal influenza advanced service follows a similar trend to the total number of flu vaccines administered. In 2021/22 the average number of flu vaccines administered per community pharmacy was 492. This is 5.92 times the average in 2015/16, which was 83.

#### **`r paste0("Figure ",7,":Cost and fees for flu vaccinations provided by contractors between 2015/16 and 2021/22")`**

```{r flu_costs_and_fees}

flu_data %>%
  dplyr::select(FINANCIAL_YEAR,flu_cost_total_csv,flu_fees_total_csv) %>%
  summarise(flu_cost_total_csv = sum(flu_cost_total_csv),
            flu_fees_total_csv = sum(flu_fees_total_csv))  %>%
  tidyr::pivot_longer(cols = starts_with("flu_"),
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::mutate(
    measure = case_when(measure == "flu_cost_total_csv" ~ "Cost of vaccines",
                        measure == "flu_fees_total_csv" ~ "Value of Fees received")
  ) %>%
  group_chart_hc(
    x = "FINANCIAL_YEAR",
    y = "values",
    group = "measure",
    type = "column",
    xLab = "Financial year",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
    hc_plotOptions(column = list(stacking = "normal"))

```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

In line with the increase in the number of flu vaccines administered each year, the costs reimbursed and fees paid to community pharmacies as part of the advanced service have increased dramatically. The costs of vaccines administered have increased by 13.3 times between 2015/16 and 2021/22, from £3.75 million to £49.8 million.

The value of fees paid to pharmacies for delivering the flu vaccination service has increased by 8.54 times between 2015/16 and 2021/22, from £5.44 million to £46.5 million.



#### New Medicine Service

The New Medicine Service (NMS) is an Advanced Service within the NHS Community Pharmacy Contractual Framework. The service provides support for people with long-term conditions newly prescribed a medicine to help improve medicines adherence; it is focused on specific patient groups and conditions. Details of the currently covered patient groups can be found in the [NHS England Service Specification](https://www.england.nhs.uk/publication/advanced-service-specification-nhs-new-medicine-service-nms/).

The fee for NMS varies between £20 and £28 depending on the volume of prescriptions dispensed by a contractor and the number of NMSs undertaken.

#### **`r paste0("Figure ",8,":Total New Medicine Services undertaken by community pharmacies between  2015/16 and 2021/22")`**

```{r advanced services nms}

national_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N") %>% 
  dplyr::select(FINANCIAL_YEAR,num_nms_total) %>%
  tidyr::pivot_longer(
    cols = c(num_nms_total),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "num_nms_total" ~ "Total New Medicine Services undertaken - community pharmacies")
  ) %>%
  group_chart_hc(
    x = "FINANCIAL_YEAR",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Financial year",
    yLab = "Total Advance Services",
    title = ""  )
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The number of NMSs claimed in 2021/22 increased by 125% from 933,000 in 2020/21 to 2.10 million in 2021/22 a return to trend after a decrease in 2020/21 following five years of increases. The number of NMSs is  155% higher than the 822,000 claimed in 2015/16. The change in trend for 2020/21 may have been due to the impact of the COVID-19 restrictions on face-to-face services between patients and pharmacies and the increase in 2021/22 may be due to a catch-up NMS service that has been introduced for those patients who were prescribed a new medicine but did not receive an NMS or any additional support during the COVID-19 pandemic. This catch-up service was offered to eligible patients between 1st September 2021 and 31st March 2022.

#### **`r paste0("Figure ",9,":Total Cost of New Medicine Services undertaken by community pharmacies between  2015/16 and 2021/22")`**


```{r advanced services nms cost}

national_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N") %>% 
  dplyr::select(FINANCIAL_YEAR,num_nms_fee_total) %>%
  tidyr::pivot_longer(
    cols = c(num_nms_fee_total),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "num_nms_fee_total" ~ "Total Cost of New Medicine Services undertaken - community pharmacies")
  ) %>%
  group_chart_hc(
    x = "FINANCIAL_YEAR",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Financial year",
    yLab = "Value (GBP)",
    title ="",
    currency = TRUE
  )

```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

In line with the increase in the number of NMSs provided, the fees paid to community pharmacies as part of the advanced service have increased by 130% from £23.4 million  in 2020/21 to £53.9 million  in 2021/22 and are 166% higher than the £20.3 million paid in 2015/16.



#### Appliance Use Review (AUR)

Appliance Use Review (AUR) is an Advanced Service in the NHS Community Pharmacy Contractual Framework (CPCF). AURs can be carried out by a pharmacist or a specialist nurse in the pharmacy or at the patient’s home.

A fee of £28 is payable for an Appliance Use Review conducted on contractor's premises or by telephone or video link where clinically appropriate <sup>1</sup> managed by the pharmacy or appliance contractor. A fee of £54 is payable for a review conducted at the patient’s home. When carrying out AURs for multiple patients living at the same property in the same 24 hours, only the first AUR is paid at the home rate with subsequent AURs paid at the lower premises rate.

The total number of Appliance Use Reviews that an appliance or pharmacy contractor may claim fees for will be limited to one for every 35 Part IXA (qualifying items), Part IXB and Part IXC prescription items dispensed in the financial year of the claim.

<small>1. This was introduced as of 1 September 2020.</small>

#### **`r paste0("Figure ",10,": Total Appliance Use Reviews (AUR) provided between 2015/16 and 2021/22")`** 

```{r advanced services aur}

national_extract %>%
  dplyr::select(FINANCIAL_YEAR, num_aur_home_total_pharm,num_aur_prem_total_pharm,num_aur_all_total_pharm
) %>%
  summarise(num_aur_home_total_pharm = sum(num_aur_home_total_pharm),
            num_aur_prem_total_pharm = sum(num_aur_prem_total_pharm),
            num_aur_all_total_pharm = sum(num_aur_all_total_pharm))  %>%
  tidyr::pivot_longer(
    cols = c(num_aur_home_total_pharm,num_aur_prem_total_pharm,num_aur_all_total_pharm
             ),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "num_aur_home_total_pharm" ~ "Total Home Appliance Use Reviews (AUR)",measure == "num_aur_prem_total_pharm" ~ "Total Premises Appliance Use Reviews (AUR)",measure == "num_aur_all_total_pharm" ~ "Total All Appliance Use Reviews (AUR)"
                        )
  ) %>%
  group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Financial year",
  yLab = "Total Advance Services",
  title = ""
)
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

In 2021/22 the number of AURs increased by 61.9% from 2020/21 which had seen a decrease after 5 years of steady increases. The change in trend for 2020/21 may have been due to the impact of the COVID-19 restrictions on face-to-face services between patients and pharmacies, though telephone and video options were introduced for AURs these were not available for the full year. The number of AURs is 92.2% higher than 2015/16.

This advanced service can be offered by both community pharmacies and appliance contractors and a breakdown by contractor type and AUR type can be found in Table 2 in the accompanying summary tables.

#### **`r paste0("Figure ",11,": Total Cost of Appliance Use Reviews (AUR) provided between 2015/16 and 2021/22")`** 
```{r advanced services aur cost}
national_extract %>%
  dplyr::select(FINANCIAL_YEAR, num_aur_home_total_fee_pharm,num_aur_prem_total_fee_pharm,num_aur_all_total_fee_pharm
) %>%
  summarise(num_aur_home_total_fee_pharm = sum(num_aur_home_total_fee_pharm),
            num_aur_prem_total_fee_pharm = sum(num_aur_prem_total_fee_pharm),
            num_aur_all_total_fee_pharm = sum(num_aur_all_total_fee_pharm))  %>%
  tidyr::pivot_longer(
    cols = c(num_aur_home_total_fee_pharm,num_aur_prem_total_fee_pharm,num_aur_all_total_fee_pharm
             ),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "num_aur_home_total_fee_pharm" ~ "Total Home Appliance Use Reviews (AUR)",measure == "num_aur_prem_total_fee_pharm" ~ "Total Premises Appliance Use Reviews (AUR)",measure == "num_aur_all_total_fee_pharm" ~ "Total All Appliance Use Reviews (AUR)"
                        )
  ) %>%
  group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Financial year",
  yLab = "Value (GBP)",
  title = "",
  currency =TRUE
)
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The total cost of fees for Appliance Use Reviews increased 30.9% in 2021/22 to £2.73 million from the £2.09 million paid in 2021/22 in line with the increase in the number of reviews undertaken,  the total paid was 47.9% higher than the £1.85 million paid in 2015/16.

#### Stoma Appliance Customisation

Stoma Appliance Customisation (SAC) is an Advanced Service in the NHS Community Pharmacy Contractual Framework (CPCF). The service involves the customisation of stoma appliances, based on the patient’s measurements or a template. The aim of the service is to ensure proper use and comfortable fitting of the stoma appliance and to improve the duration of usage, thereby reducing waste. The stoma appliances that can be customised are listed in Part IXC of the Drug Tariff.

A fee of £4.32 is payable for every Part IXC prescription item that can be customised as denoted in the Drug Tariff. The fee does not have to be claimed and will be paid automatically to those pharmacy and appliance contractors  in England who have registered their intent to provide stoma appliance customisation services to the NHSBSA.

#### **`r paste0("Figure ",12,": Total  Stoma Appliance Customisation (SAC) provided between 2015/16 and 2021/22")`** 

```{r advanced services sac}
national_extract %>%
  dplyr::mutate(CONTRACTOR_TYPE =  case_when(APPLIANCE_DISPENSER_HIST == "Y" ~ "Total Stoma Appliance Customisation (SAC) provided by appliance contractors",
                                  APPLIANCE_DISPENSER_HIST == "N" ~ "Total Stoma Appliance Customisation (SAC) provided by community pharmacies")) %>% 
  dplyr::select(FINANCIAL_YEAR,num_stoma_total_pharm, CONTRACTOR_TYPE) %>% 
   group_by(FINANCIAL_YEAR, CONTRACTOR_TYPE) %>% 
  
  tidyr::pivot_longer(
    cols = c(num_stoma_total_pharm),
    names_to = "measure",
    values_to = "values") %>%
    group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "CONTRACTOR_TYPE",
  type = "line",
  xLab = "Financial year",
  yLab = "Total Advance Services",
  title = "")
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)


The number of SACs provided by community pharmacies decreased by 28.0% from 11,600 to 8,330 between 2020/21 and 2021/22. This followed 5 consecutive years of decreases between 2015/16 and 2020/21. The number of SACs provided by appliance contractors between 2020/21 and 2021/22  went up by 5.23% from 1.52 million to 1.60 million. This followed 5 consecutive years of increases between 2015/16 and 2021/22.

The majority of SACs are provided by appliance contractors, which means the reduction in those provided by community pharmacies has not prevented an overall increase. The total number of SACs has increased by 4.98% from 1.54 million to 1.61 million between 2020/21 and 2021/22. This is an increase of 30.2% from the 1.24 million provided in 2015/16.

#### **`r paste0("Figure ",13,": Total  Stoma Appliance Customisation (SAC) provided between 2015/16 and 2021/22")`** 
```{r advanced services sac cost}

national_extract %>%
  dplyr::mutate(CONTRACTOR_TYPE =  case_when(APPLIANCE_DISPENSER_HIST == "Y" ~ "Total Cost of Stoma Appliance Customisation (SAC) provided by appliance contractors",
                                  APPLIANCE_DISPENSER_HIST == "N" ~ "Total Cost of Stoma Appliance Customisation (SAC) provided by community pharmacies")) %>% 
  dplyr::select(FINANCIAL_YEAR,num_stoma_total_fee_pharm, CONTRACTOR_TYPE) %>% 
   group_by(FINANCIAL_YEAR, CONTRACTOR_TYPE) %>% 
  
  tidyr::pivot_longer(
    cols = c(num_stoma_total_fee_pharm),
    names_to = "measure",
    values_to = "values") %>%
    group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "CONTRACTOR_TYPE",
  type = "line",
  xLab = "Financial year",
  yLab = "Value (GBP)",
  title = "",
  currency = TRUE)
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The fee for Stoma Appliance Customisation is a set payment that has not changed since its introduction so the increase in the cost of the fees  mirrors the trends for number carried out. £6.96 million  was spent in total in 2021/22 compared to £6.63 million in 2015/16. In line with the increase in the number of SACs provided each year, the fees paid to community pharmacies as part of the advanced service have increased. As the SAC is a set payment that has not changed since its introduction this percentage increase for 2021/22 is also 4.98%.

####  Community Pharmacist Consultation Services

The NHS Community Pharmacist Consultation Service launched on 29th October 2019 as an Advanced Service.

Since 1st November 2020, general practices have been able to refer patients for a minor illness consultation via CPCS, once a local referral pathway has been agreed.

The fee for providing the Community Pharmacist Consultation Service is £14 (including VAT) per delivery.

<!-- #### **`r paste0("Figure ",16,": Total Community Pharmacy Consultation Services 2021/22")`** -->

<!-- ```{r cspc num} -->
<!-- national_extract %>% -->
<!--   filter(APPLIANCE_DISPENSER_HIST == "N", -->
<!--          FINANCIAL_YEAR %in% c("2019/2020", "2020/2021","2021/2022")) %>%  -->
<!--   dplyr::select(FINANCIAL_YEAR, cpcs_fees_pharm) %>% -->
<!--   dplyr::mutate(cpcs_fees_pharm_count = round(cpcs_fees_pharm/14,0)) %>% -->
<!--   tidyr::pivot_longer( -->
<!--     cols = c(cpcs_fees_pharm_count), -->
<!--     names_to = "measure", -->
<!--     values_to = "values" -->
<!--   ) %>% -->
<!--    dplyr::mutate( -->
<!--     measure = case_when(measure == "cpcs_fees_pharm_count" ~ "Total Community Pharmacy Consultation Services") -->
<!--   ) %>% -->
<!--   group_chart_hc( -->
<!--   x = "FINANCIAL_YEAR", -->
<!--   y = "values", -->
<!--   group = "measure", -->
<!--   type = "line", -->
<!--   xLab = "Financial Year", -->
<!--   yLab = "Total Advance Services", -->
<!--   title = "" -->
<!-- )   -->
<!-- ``` -->
<!-- Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx) -->

No data prior to 2019/20 is available for Community Pharmacy Consultation Services. A total of 896,000 Community Pharmacist Consultation Service and 136% increase on the 380,000 claimed in 2020/21.

#### **`r paste0("Figure ",14,": Total Cost of Community Pharmacy Consultation Services 2021/22")`**
```{r cspc fee cost}
national_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N",
         FINANCIAL_YEAR %in% c("2019/2020", "2020/2021","2021/2022")) %>% 
  dplyr::select(FINANCIAL_YEAR, cpcs_fees_pharm) %>%
  tidyr::pivot_longer(
    cols = c(cpcs_fees_pharm),
    names_to = "measure",
    values_to = "values"
  ) %>%
   dplyr::mutate(
    measure = case_when(measure == "cpcs_fees_pharm" ~ "Total Cost of Community Pharmacy Consultation Services")
  ) %>%
  group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Financial Year",
  yLab = "Value (GBP)",
    title ="",
    currency = TRUE) 

```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of fees paid for Community Pharmacy Consultation Services in 2021/22 was £12.5 million, a 81.4% increase on the £5.31 million  claimed in 2020/21.

#### **`r paste0("Figure ",15,": Total Cost of GP referral pathway to the Community Pharmacist Consultation Service engagement fees")`**
```{r cspc drug cost}
national_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N",
         FINANCIAL_YEAR %in% c("2020/2021","2021/2022")) %>% 
  dplyr::select(FINANCIAL_YEAR, GPREFPATH_FEES) %>%
  tidyr::pivot_longer(
    cols = c(GPREFPATH_FEES),
    names_to = "measure",
    values_to = "values"
  ) %>%
   dplyr::mutate(
    measure = case_when(measure == "GPREFPATH_FEES" ~ "Total Cost of GP referral pathway engagement fees")
  ) %>%
  group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Financial Year",
  yLab = "Value (GBP)",
    title ="",
    currency = TRUE) 

```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

From November 2020, pharmacies that completed engagement activity in  the NHS CPCS service specification could also receive a £300 activity payment called the GP referral pathway engagement fee.

The cost of GP referral pathway engagement fees in 2021/22 was £1.76 million, a 39.2 increase on the £1.27 million  claimed in 2020/21.

#### **`r paste0("Figure ",16,": Total Cost of Drugs provided during Community Pharmacy Consultation Services 2021/22")`**
```{r cspc drug cost}
national_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N",
         FINANCIAL_YEAR %in% c("2019/2020", "2020/2021","2021/2022")) %>% 
  dplyr::select(FINANCIAL_YEAR, cpcs_drugs_pharm) %>%
  tidyr::pivot_longer(
    cols = c(cpcs_drugs_pharm),
    names_to = "measure",
    values_to = "values"
  ) %>%
   dplyr::mutate(
    measure = case_when(measure == "cpcs_drugs_pharm" ~ "Total Cost of Drugs provided during Community Pharmacy Consultation Services")
  ) %>%
  group_chart_hc(
  x = "FINANCIAL_YEAR",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Financial Year",
  yLab = "Value (GBP)",
    title ="",
    currency = TRUE) 

```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of drugs paid for Community Pharmacy Consultation Services in 2021/22 was £2.84 million, a 136% increase on the £1.56 million  claimed in 2020/21.


#### Hepatitis C testing service

Hepatitis C testing service is an Advanced Service in the NHS Community Pharmacy Contractual Framework (CPCF). The service offers people who inject drugs(PWIDs), who are not engaged with community drug and alcohol treatment services, opportunity to receive a point of care testing (POCT)  HCV test from a community pharmacy.

Pharmacies  are  able to claim a payment of £36 per test performed on an eligible PWID, plus the cost of the test kit at the manufacturer’s list price (including VAT).

#### **`r paste0("Figure ",17,":Total Costs for Hepatitis C testing service fees provided by contractors in 2021/22")`**

```{r hep_c_service_pharm}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,hep_c_service_pharm) %>%
  summarise(hep_c_service_pharm = sum(hep_c_service_pharm))  %>%
  tidyr::pivot_longer(cols = hep_c_service_pharm,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "hep_c_service_pharm" ~ "Cost of Hepatitis C testing service fees")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of Hepatitis C testing service fees administered by community pharmacies in 2021/22 was £10,000.

#### **`r paste0("Figure ",18,":Total Costs for Hepatitis C testing kit provided by contractors in 2021/22")`**

```{r hep_c_kit_pharm}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,hep_c_kit_pharm) %>%
  summarise(hep_c_kit_pharm = sum(hep_c_kit_pharm))  %>%
  tidyr::pivot_longer(cols = hep_c_kit_pharm,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "hep_c_kit_pharm" ~ "Cost of Hepatitis C testing kits")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of Hepatitis C testing service fees administered by community pharmacies in 2021/22 was £1,670.


#### Hypertension Case-Finding Advanced Service

Hypertension Case-Finding is an Advanced Service in the NHS Community Pharmacy Contractual Framework (CPCF). The service was introduced in October 2021 following a NHS England pilot where pharmacies offered blood pressure checks to people 40 years and over and 24 hour ambulatory blood pressure monitoring (ABPM) to some patient's with elevated initial blood pressure readings.


Pharmacy contractors who have signed up to deliver the NHS Community Pharmacy Hypertension Case-Finding Advanced Service qualify for a set up payment of £440 to support initial costs incurred in setting up the service.

A fee of £15 will be paid for each patient receiving a clinic blood pressure check. A fee of £45 will be paid for ABPM provided to a patient. 

Additionally, pharmacy contractors who deliver the following minimum number of ABPM checks each year will automatically trigger an incentive payment. For 2021/22 the minimum number of ABPM checks is five and the incentive payment is £1,000.

#### **`r paste0("Figure ",19,":Total Costs for Hypertension Case-Finding Advanced Service fees provided by contractors in 2021/22")`**

```{r HYPTENCHECK_FEES}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,HYPTENCHECK_FEES) %>%
  summarise(HYPTENCHECK_FEES = sum(HYPTENCHECK_FEES))  %>%
  tidyr::pivot_longer(cols = HYPTENCHECK_FEES,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "hep_c_service_pharm" ~ "Cost of Hypertension Case-Finding Advanced Service fees")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of Hypertension Case-Finding Advanced Service fees in 2021/22 was £3.42 million.

#### **`r paste0("Figure ",20,":Total Costs for Hypertension Case-Finding Advanced Service Set up payments provided by contractors in 2021/22")`**

```{r HYPTENSET_FEES}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,HYPTENSET_FEES) %>%
  summarise(HYPTENSET_FEES = sum(HYPTENSET_FEES))  %>%
  tidyr::pivot_longer(cols = HYPTENSET_FEES,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "HYPTENSET_FEES" ~ "Cost of Hypertension Case-Finding Advanced Service Set up payments")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of Hypertension Case-Finding Advanced Service Set up payments in 2021/22 was £4.03 million.

#### **`r paste0("Figure ",21,":Total Costs for Hypertension Case-Finding Advanced Service Incentive up payments provided by contractors in 2021/22")`**

```{r HYPTEN_INC}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,HYPTEN_INC) %>%
  summarise(HYPTEN_INC = sum(HYPTEN_INC))  %>%
  tidyr::pivot_longer(cols = HYPTEN_INC,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "HYPTEN_INC" ~ "Cost of Hypertension Case-Finding Advanced Service Incentive payments")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of Hypertension Case-Finding Advanced Service Incentive up payments in 2021/22 was £713,000

#### Community Pharmacy Home Delivery Services

The Community Pharmacy Home Delivery Service was introduced during the COVID-19 pandemic. Initially used to deliver prescriptions to  patients who were extremely clinically vulnerable the service was active from April 2020 to 31 July 2020, then from 5 November 2020 to 3 December 2020. The service continued in Tier 4 areas until it was recommenced from 5 January 2021 to 31 March 2021. After this date the service was ended for  clinically extremely vulnerable (CEV) patients but is still available to patients who have been advised to self-isolate by NHS Test and Trace. Home Delivery Services became available to self-isolating patients on 16 March 2021. The data held does not differentiate between delivery services for CEV and those in self-isolation.

The fee for providing the Home Delivery Service is £6 (including VAT) per delivery.

#### **`r paste0("Figure ",22,": Total Community Pharmacy Home Delivery Services 2021/22")`**

```{r advanced services home del}
national_month_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N") %>% 
  dplyr::select(YEAR_MONTH, num_home_del_total) %>%
  tidyr::pivot_longer(
    cols = c(num_home_del_total),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "num_home_del_total" ~ "Total Community Pharmacy Home Delivery Services")
  ) %>%
  group_chart_hc(
  x = "YEAR_MONTH",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Month",
  yLab = "Total Advance Services",
  title = ""
)  %>%
  hc_xAxis(type = "datetime")
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

No data prior to 2020/21 is available for Community Pharmacy Home Delivery Services which were introduced during the COVID-19 pandemic; therefore, we have presented this data monthly. Variation in the monthly amount of this service lines up with the time periods it was active. A total of 456,000 Home Delivery Services were claimed with the greatest number being in April 2021 with 135,000. This service was ended on 31 March 2022;

#### **`r paste0("Figure ",23,": Total Cost of Community Pharmacy Home Delivery Services 2021/22")`**
```{r advanced services home cost}
national_month_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N") %>% 
  dplyr::select(YEAR_MONTH, home_del_cost_total) %>%
  tidyr::pivot_longer(
    cols = c(home_del_cost_total),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "home_del_cost_total" ~ "Total Cost of Community Pharmacy Home Delivery Services")
  ) %>%
  group_chart_hc(
  x = "YEAR_MONTH",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Month",
  yLab = "Value (GBP)",
    title ="",
    currency = TRUE) %>% 
#hc_tooltip(enabled = T,             shared = T,             sort = T) %>% 
  hc_xAxis(type = "datetime")
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of fees paid for the Home Delivery Services had the same variation as number of home deliveries with a total of £2.74 million paid for Home Delivery Services. April 2021 was the highest value month with £808,000 in fees paid.

#### Community Pharmacy Covid-19 Lateral Flow Device Distribution Service

Community Pharmacy Covid-19 Lateral Flow Device Distribution Service is an Advanced Service in the NHS Community Pharmacy Contractual Framework (CPCF). The service was introduced in March 2021

The service made lateral flow device (LFD) test kits readily available at community pharmacies. The service was part of the Government’s offer of lateral flow testing to all people in England and it worked alongside NHS Test and Trace’s other COVID-19 testing routes

Pharmacies who registered for the service received a one-off set-up fee of £250 + VAT. An early sign-up fee of £200 + VAT, was also available for pharmacies who signed up to provide the service by 23:59 on 18th April 2021.

From 29 March 2021, a service fee of £1.50 + VAT per transaction applied. 

From 4 October 2021 there were two service fees: 

* A service fee of £1.70 + VAT per transaction using a collect code
* A service fee of £1.20 + VAT per transaction not using a collect code.

#### **`r paste0("Figure ",24,": Total Cost of ommunity Pharmacy Covid-19 Lateral Flow Device Distribution Services fees 2021/22")`**
```{r advanced services home cost}
national_month_extract %>%
  filter(APPLIANCE_DISPENSER_HIST == "N") %>% 
  dplyr::select(YEAR_MONTH,CVDTEST_KITS) %>%
  tidyr::pivot_longer(
    cols = c(CVDTEST_KITS),
    names_to = "measure",
    values_to = "values"
  ) %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "CVDTEST_KITS" ~ "Total Cost of Community Pharmacy Covid-19 Lateral Flow Device Distribution Service fees")
  ) %>%
  group_chart_hc(
  x = "YEAR_MONTH",
  y = "values",
  group = "measure",
  type = "line",
  xLab = "Month",
  yLab = "Value (GBP)",
    title ="",
    currency = TRUE) %>% 
#hc_tooltip(enabled = T,             shared = T,             sort = T) %>% 
  hc_xAxis(type = "datetime")
```
Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The cost of fees paid for the Home Delivery Services had the same variation as number of home deliveries with a total of £2.74 million paid for Home Delivery Services. April 2021 was the highest value month with £808,000 million in fees paid.


<!-- #### PPE Claims  -->

<!-- From 1 January 2021, pharmacy contractors were able to claim for PPE that was purchased between 27 February 2020 and 31 December 2020 for use in the delivery of NHS pharmaceutical services up to 31 March 2021 as a result of COVID-19 infection control guidance. -->

<!-- Contractors could order free of charge PPE for use in their pharmacy via the DHSC’s PPE portal from 1 January 2021. -->

<!-- £20,200  -->


<!-- #### Reimbursement for COVID-19 costs -->

<!-- Pharmacy contractors were able to submit claims for reimbursement of COVID-19 related costs incurred between 1 March 2020 and 31 March 2021. Contractors were able to submit claims for: -->

<!-- *additional staff costs due to COVID-19 -->
<!-- *costs for making your premises COVID-secure -->
<!-- *IT and communication costs to support remote working and virtual patient contact due to COVID-19 -->
<!-- *closures agreed by NHS England and NHS Improvement -->

<!-- All claims had to be  submitted to the NHSBSA between 5 July and 15 August 2021 -->
<!-- <!-- #### **`r paste0("Figure ",21,":Total Cost for COVID-19  costs provided by contractors in 2021/22")`** --> -->

<!-- <!-- ```{r covid__costs} --> -->

<!-- <!-- national_month_extract %>% --> -->
<!-- <!--   dplyr::select(YEAR_MONTH,CVD_19_COSTS) %>% --> -->
<!-- <!--   summarise(CVD_19_COSTS = sum(CVD_19_COSTS))  %>% --> -->
<!-- <!--   tidyr::pivot_longer(cols = CVD_19_COSTS, --> -->
<!-- <!--                       names_to = "measure", --> -->
<!-- <!--                       values_to = "values") %>% --> -->
<!-- <!--   dplyr::arrange(desc(measure)) %>% --> -->
<!-- <!--   mutate( --> -->
<!-- <!--     YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d") --> -->
<!-- <!--   ) %>% --> -->
<!-- <!--   dplyr::mutate( --> -->
<!-- <!--     measure = case_when(measure == "CVD_19_COSTS" ~ "COVID-19 Costs") --> -->
<!-- <!--   ) %>% --> -->
<!-- <!--   group_chart_hc( --> -->
<!-- <!--     x = "YEAR_MONTH", --> -->
<!-- <!--     y = "values", --> -->
<!-- <!--     group = "measure", --> -->
<!-- <!--     type = "line", --> -->
<!-- <!--     xLab = "Month", --> -->
<!-- <!--     yLab = "Value (GBP)", --> -->
<!-- <!--     title = "", --> -->
<!-- <!--     currency = TRUE --> -->
<!-- <!--   ) %>% --> -->
<!-- <!--   hc_xAxis(type = "datetime") --> -->
<!-- <!-- ``` --> -->

<!-- <!-- Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx) --> 

<!-- The additional costs of COVID-19 claimed  by community pharmacies in 2021/22 were £264 million. -->


### Enhanced Services for Contractors

#### COVID-19  vaccination enhanced service

Towards the end of 2020/21 community pharmacies began providing COVID-19 vaccinations as a local enhanced service to deliver Phase 1 and Phase 2 cohorts  of the Joint Committee on Vaccination and Immunisation (JCVI). A further enhanced service was commissioned to deliver Phase 3 which was provided alongside the seasonal influenza vaccination programme, which runs each autumn.

The figures in these statistics for COVID-19 vaccinations only relate to those administered by community pharmacies. [Official Statistics for these are produced by UK Health Security Agency England](https://coronavirus.data.gov.uk/).

#### **`r paste0("Figure ",20,":Total Costs for COVID-19  vaccinations fees provided by contractors in 2021/22")`**

```{r covidvaccine_costs_and_fees}

national_month_extract %>%
  dplyr::select(YEAR_MONTH,CVD_19_VACCINE) %>%
  summarise(CVD_19_VACCINE = sum(CVD_19_VACCINE))  %>%
  tidyr::pivot_longer(cols = CVD_19_VACCINE,
                      names_to = "measure",
                      values_to = "values") %>%
  dplyr::arrange(desc(measure)) %>%
  mutate(
    YEAR_MONTH = base::as.Date(as.character(paste0(YEAR_MONTH,"01")), format = "%Y%m%d")
  ) %>%
  dplyr::mutate(
    measure = case_when(measure == "CVD_19_VACCINE" ~ "Cost of vaccines")
  ) %>%
  group_chart_hc(
    x = "YEAR_MONTH",
    y = "values",
    group = "measure",
    type = "line",
    xLab = "Month",
    yLab = "Value (GBP)",
    title = "",
    currency = TRUE
  ) %>%
  hc_xAxis(type = "datetime")
```

Source: [Summary Tables - General Pharmaceutical Services - Table 2](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-summary-tables-v001.xlsx)

The costs of COVID-19 vaccines fees in 2021/22 were £334 million, with a peak in Autumn when Phase 3 launched.


---

## 3. About these statistics

The General Pharmaceutical Services publication (GPhS) is a National Statistic release. National Statistics Status means that GPhS meets the highest standards of trustworthiness, quality, public value, and complies with all aspects of the [Code of Practice for Statistics](https://code.statisticsauthority.gov.uk/).

The designation of this publication as a National Statistic was confirmed in December 2021 following an assessment by the Office for Statistics Regulation (OSR).



### 3.1. Changes made to this publication {.toc-ignore}

Table 1 and Table 2 in previous releases has been split into various tables to make the tables more user-friendly. Data around the number of contractors is now grouped with the activity by contractors and split by service type.

Data regarding other COVID-19 related services that have were undertaken by pharmacies are included in the summary tables in this release.

This includes data and payments on:

* COVID-19 personal protective equipment (PPE).
* COVID-19 premises and refrigeration set up.
* COVID-19 testing.
* COVID-19 vaccines.


Data regarding the essential service Discharge Medicine Service and the Advanced Service Hypertension Case-Finding Advanced Service have also been added.

Previously some figures in the supplementary tables regarding fees and fee counts had been rounded to the nearest 10 to reflect that these figures are estimated based on claims by contractors and that work will be done to confirm these payments. Users should always use the statistics in the latest publication to make sure they are the most up to date figures available.

Changes to the NHS Organisational Structure mean that Integrated care boards (ICBs) will succeed sustainability and transformation plans (STPs) from July 2022 which is reflected in the summary tables.


<!-- ###  Changes to Methodology -->

<!-- The first release of this publication by the NHS Business Services Authority published in November 2020 had a number of methodology changes from previous General Pharmaceutical Services publications by NHS Digital. -->

<!-- These methodologies have subsequently been used for this publication and include: -->

<!--   * __Counts of community pharmacies and appliance contractors.__ -->
<!-- These counts are  based around contractors that have been active at __any__ point in the given year. This has been done to provide consistency   with other figures given in the publication that do not exclude contractors that have  closed during the year. Previous releases of these       statistics by NHS Digital counted active contractors as those open on 31 March of the given year.  This change avoids the exclusion of          contractors that have opened and closed in the same financial year. -->
<!-- * __Average monthly items per contractor.__ -->
<!-- This measure is calculated by, for each pharmacy, dividing the total items dispensed by the number of months the pharmacy was active in the     year. The median of these figures is then calculated to give the final measure. A median is calculated by arranging all the available values    into an ordered list and selecting the value that is in the middle. If there are 2 middle values, the median is halfway between them.           Previously this measure was calculated by dividing the total number of items dispensed by community pharmacies by the total number of           community pharmacies, and then dividing this number by the number of months in the year (total items dispensed / number of community            pharmacies / 12). This change means the average is no longer skewed by contractors who are not open for all months of the year or by outliers   in the volume of dispensed items. -->
<!-- * __Contractor attributes are all taken from the 31 March of the year.__ -->
<!-- Therefore, if a community pharmacy has a change in status in March, this is the attribute that would be used.  This differs from -->
<!-- historical releases by NHS Digital where a single contractor could be counted multiple times if an attribute was changed during the financial   year. For example a contractor type could move from independent to multiple if another branch or branches are opened. -->
<!-- * __Seasonal influenza vaccination advanced service.__ -->
<!-- The count of community pharmacies providing this service has been done on the pharmacy’s dispenser code, which remains the same if a            contractor changes trading address, to remove duplication and obtain a more accurate count.  This differs from  historical releases by NHS      Digital where the count was done on trading address meaning a   contractor could be counted multiple times if its trading address changed       during the financial year. -->
<!-- * __Monthly dispensing volume bands__. -->
<!-- Monthly volume bands are calculated using the months that a contractor has been active throughout the year. This has been done to avoid the data being skewed for contractors that opened or closed during the year. Previously this measure was calculated by the number of months in the year. This change means data won't be skewed by contractors who are not open for all months of the year. -->

<!-- Further information on the methodology used in this publication and more background information is available in the [Background Information and Methodology](https://www.nhsbsa.nhs.uk/sites/default/files/2021-10/gps_2021_background_information_methodology_v001.pdf) supporting document that accompanies this release. -->

### 3.2. Accuracy of these statistics {.toc-ignore}
 {.toc-ignore}

Further information on the accuracy of the data used in this publication and the impact of post payment verification (PPV) and other checks carried out by NHS Prescription Services (a division of NHSBSA) on the statistics is available in the [Impacts of errors identified on the accuracy of the General Pharmaceutical Services](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-accuracy-v001.html) supporting document that accompanies this release.

### 3.3. PHS1 Pharmacy Data Collection  {.toc-ignore}

Data that has previously been obtained from the NHS Digital PHS1 Pharmacy Data collection is no longer available, as the collection exercise no longer takes place. An alternative data source for this data has not been identified which means this information has not been included in this report. However, some information previously included on the PHS1, such as pharmacy openings and closures and some exempt categories has been sourced from the NHSBSA administrative records.


Versions of this publication released by NHS Digital are [available on the NHS Digital website](https://digital.nhs.uk/data-and-information/publications/statistical/general-pharmaceutical-services). Users should always use the statistics in the latest publication to make sure they are the most up to date figures available.

### 3.4. Uses of these statistics {.toc-ignore}

These statistics can be used to understand the activity of community pharmacies and appliance contractors in England, including essential and advanced services provided to NHS patients. They cannot be used to provide the total number or cost of prescription items dispensed in England in the community as they do not contain dispensing from all sources. This can be obtained from another NHSBSA National Statistic publication, [Prescription Cost Analysis](https://www.nhsbsa.nhs.uk/statistical-collections/prescription-cost-analysis-england).


### 3.5. Timeliness of this publication {.toc-ignore}

The publication date by NHS Digital has historically been in November, as continued by the first in series of this publication by the NHSBSA. The publication date has been brought forward to October for this publication and the NHSBSA will be looking to bring this forward again for the 2022/23 release.

### 3.6. Geographies included in this publication {.toc-ignore}

The geographies used in this publication are based upon NHSBSA administrative records, not geographical health boundaries as defined by the Office for National Statistics (ONS). These administrative records more closely reflect the operational organisation of dispensing contractors than other geographical data sources such as the National Statistics Postcode Lookup (NSPL).

The NHS England Regions and Integrated care boards (ICBs)  shown in the statistical summary tables of this release are based on this new NHS organisational structure as of July 2022

<!-- ### Planned changes to this publication {.toc-ignore} -->



<!-- The NHSBSA is currently developing an open data set public feed based on the monthly [Management Information Spreadsheet](https://www.nhsbsa.nhs.uk/access-our-data-products/information-services-portal-isp/isp-report-information) (MIS) report. This is intended for release by March 2022 and will compliment this National Statistic publication.  This new data set will not be an Official Statistic release. -->

<!-- Following the release of this publication the NHSBSA will be launching a public consultation on the statistical methodology that is used in these statistics and will be seeking feedback from users and stakeholders. NHSBSA also plan to hold a user engagement event for this release in early 2021. Details of the event will be announced on the [Statistics section](https://www.nhsbsa.nhs.uk/announcements) of the NHSBSA website. -->

<!-- NHSBSA plans to expand the scope of this publication in future releases to report on all fees that a contractor can claim for providing NHS services in order to provide a more complete picture of the total cost of community pharmacy to the public purse. -->

---

## 4. Statistical Disclosure Control

The high-level figures in this statistical summary have been rounded where appropriate for clarity, in most cases to three significant figures. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Some figures in the supplementary tables regarding fees and fee counts have been rounded to the nearest 10 to better reflect that these figures are estimated based on claims by contractors and that work will be done to confirm these payments.

---

## 5. Accessibility

### How to request content in an accessible format

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: nhsbsa.accessibility@nhs.net

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---


## 6. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](https://online1.snapsurveys.com/Official_Statistics_Feedback) to help us improve the Official Statistics that we produce. All responses will remain anonymous, and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** nhsbsa.statistics@nhs.net

**You can also write to us at:**

NHSBSA - Statistics<br>
NHS Business Services Authority<br>
Stella House<br>
Goldcrest Way<br>
Newburn Riverside<br>
Newcastle upon Tyne<br>
NE15 8NY


**Responsible statistician:** Kirsty Gray


</main>
