---
title: "`r config$publication_top_name`"
pagetitle: GPhS England
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '2'
  html_document:
    anchor_sections: no
    css: 
      - www/style.css
      - www/nhs-min.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>

  
    $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/Accredited Official Statistics Logo English.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%; margin-top:140px;\">');
  });
</script>

<html lang="en">

```{r setup, include=FALSE}

# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)


```

<div class = "quartoheader", id = "fixed-header">
        <header class="nhsuk-header" role="banner">
    <div class="nhsuk-width-container nhsuk-header__container">
      <div class="nhsuk-header__logo nhsuk-header__logo--only">
        <a class="nhsuk-header__link" href="https://www.nhsbsa.nhs.uk/" aria-label="NHSBSA home" target="_blank">
          <img class="nhsuk-logo" src="www/logo-nhsbsa.svg"" name="NHSBSA logo" alt="NHS Business Services Authority">
        </a>
      </div>
    </div>
    </div>

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 




## Key findings

In England in 2023/24 for community pharmacies and appliance contractors:

* there were 12,009 active community pharmacies and 112 active appliance contractors.For important information about interpreting this figure please see [section 3](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gphs_annual_2023_24_v001.html#about-these-statistics)

* 1.1 billion prescription items were dispensed by community pharmacies. This was a 3% increase from 2022/23

* 12 million prescription items were dispensed by appliance contractors. This was a 5% increase from 2022/23

* 1.08 billion prescription items were dispensed via the Electronic Prescription Service (EPS). This was 96% of all items dispensed, and a small percentage increase from 2022/23

* the cost of drugs and appliances reimbursed totalled £10 billion. This was a 5% increase from 2022/23



---

## 1. Things you should know

### 1.1 Introduction {.toc-ignore} 



This publication is an accredited official statistic release. Accredited official statistics status means that General Pharmaceutical Services (GPhS) - England complies with the [Code of Practice for Statistics](https://code.statisticsauthority.gov.uk/), showing trustworthiness, quality and value. 
    
The designation of this publication as an accredited official statistic was confirmed in December 2021 following an independent review by the Office for Statistics Regulation (OSR). You can read more about this confirmation and the assessment of these statistics on the [OSR website.](https://osr.statisticsauthority.gov.uk/correspondence/ed-humpherson-to-matthew-wilson-confirmation-of-national-statistics-designation-for-statistics-on-general-pharmaceutical-services-in-england/)

### 1.2 Background {.toc-ignore} 


From 1 April 2013, NHS England has been responsible for the commissioning of NHS pharmaceutical services in England and for negotiating changes to arrangements for the provision of services. The [Community Pharmacy Contractual Framework](https://cpe.org.uk/quality-and-regulations/the-pharmacy-contract/) (CPCF) for pharmacy contractors is set out in [The National Health Service (Pharmaceutical and Local Pharmaceutical Services) Regulations 2013](https://www.legislation.gov.uk/uksi/2013/349/contents/made), and the Pharmaceutical Services (Advanced and Enhanced Services) (England) Directions 2013 are contained within the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff).

The 2013 regulations state that a person who is not a doctor or dentist must be included in a pharmaceutical list in order to be permitted to provide pharmaceutical services for the NHS. NHS England are required to prepare and maintain lists of those whose applications have been granted to be listed as a pharmaceutical provider for both community pharmacy and appliance contractors. The lists specify both the premises and the named contractor.

To receive payment for the costs and fees incurred while providing services to the general public on behalf of the NHS, community pharmacies and appliance contractors must submit their prescriptions to NHSBSA along with a submission document, known as the FP34C. This has recently been digitised by the Manage Your Submission (MYS) application but was historically done via a paper form that accompanied paper prescription batches.

Data on advanced services such as the provision of New Medicines Service (NMS) and Appliance Use Reviews (AURs) is taken from the FP34C submission document, or from the MYS application, and passed to NHSBSA payment systems.  There it is used to calculate payments that are to be made to community pharmacies and appliance contractors

### 1.3 Scope {.toc-ignore} 

Data is collected from prescriptions submitted to the NHSBSA by dispensing contractors (community pharmacies and appliance contractors). These prescriptions are issued by GPs or other authorised prescribers such as nurses, dentists, and allied health professionals. Prescriptions that are issued by hospitals may also be dispensed in the community and submitted for reimbursement. Prescriptions that are issued in hospitals and fulfilled by the hospital pharmacy or dispensary are not included in this data.

Data is limited in this publication to only prescription items that have been dispensed by a community pharmacy or appliance contractor in England. Items dispensed by dispensing doctors, hospitals or prisons or submitted for reimbursement via a personal administration account have been excluded.

Data on advanced services provided by community pharmacies and appliance contractors is limited to where a claim has been submitted to NHSBSA in relation to performing one of those services. Counts of the number of services provided are based upon the number of times a fee has been paid to a contractor. Figures for the number of services provided are counts of the number of times a fee has been paid to a contractor for that service.


This publication aims to describe the activity of community pharmacies and appliance contractors in England across the whole financial year.

These statistics detail:

* the number of prescription items dispensed
* the reimbursement costs of prescription items dispensed
* the number of single activity fees  received
* the remuneration costs of single activity fees received
* the number of essential and advanced services provided
* the associated costs for essential and advanced services

These statistics cover the financial years between April 2015 and March 2024.


### 1.4 Definitions {.toc-ignore} 
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Item", text = "A single unit of medication listed separately on a prescription form. In this publication, an example of an item would be Fluoxetine 20mg tables x56.", width = "100%")`
::: 
::::
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Dispensing contractor", text = "A dispensing contractor is either a  community pharmacy or appliance contractor. Community pharmacies can dispense both drugs and appliances, but appliance contractors are limited to the supply of appliances as listed in Part IXA, IXB and IXC of the Drug Tariff.", width = "100%")`
::: 
::::
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Form", text = "Prescription forms are either paper forms or electronic messages sent via the Electronic Prescription Service (EPS). EPS allows prescribers to send prescriptions electronically to a dispenser, such as a pharmacy, of the patient's choice. EPS makes the prescribing and dispensing process more efficient and convenient for patients and staff.", width = "100%")`
::: 
::::
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}   
`r infoBox_border("Drug Cost", text = "There are many costs incurred when a dispensing contractor fulfils a prescription. The drug cost refers to the basic price of the item and quantity prescribed that is reimbursed back to the contractor. This is sometimes called the 'Net Ingredient Cost' (NIC). Costs paid to contractors to cover the cost of dispensed items are referred to as reimbursement. The basic price is determined by the [Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff) or by the manufacturer, wholesaler, or supplier of the product.", width = "100%")`
::: 
::::
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}   
`r infoBox_border("Fees", text = "Fees are payable to dispensing contractors for dispensing prescriptions and for providing essential and advanced services to NHS patients.  Many, but not all, fees are described in these statistics and are included in the accompanying summary tables. Details of what fees are payable to community pharmacies and appliance contractors can be found in the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff). Fee payments to dispensing contractors for prescribing and other services are known as remuneration.", width = "100%")`
::: 
::::  
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}   
`r infoBox_border("Essential services", text = "Essential services are those listed in the NHS CPCF that all pharmacy contractors must provide to NHS patients.", width = "100%")`
::: 
:::: 
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}   
`r infoBox_border("Advanced services", text = "There are several advanced services within the NHS Community Pharmacy Contractual Framework (CPCF). Community pharmacies can choose to provide any of these services as long as they meet the requirements set out in the Secretary of State Directions.", width = "100%")`
::: 
:::: 
::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}   
`r infoBox_border("Enhanced services", text = "As well as national services provided by all pharmacies, the NHS Community Pharmacy Contractual Framework also includes services that are commissioned at a local level. These are known as Local Enhanced Services (LES).
In December 2021, provisions were made within the NHS (Pharmaceutical and Local Pharmaceutical Services) Regulations 2013 for a new type of Enhanced service, the National Enhanced Service. Under this type of service, NHS England commissions an enhanced service that is nationally specified.", width = "100%")`
::: 
:::: 

Further definitions of terms used in this publication can be found in the glossary of the [Background Information and Methodology](`r config$background_link` ) supporting document that accompanies this release.

---

## 2. Results and commentary

The charts in these statistics are interactive and allow you to isolate the aspect that you are interested in by selecting or deselecting a series from the legend of the chart. For example, if you only want to view data for community pharmacies, you can click the other categories in the legend of the chart to deselect them. If you wish to include them again just click on their legend entry once more.

::::{.row style="display: flex; margin-bottom: 24px;"} 
:::{.col-md-12} 
`r infoBox_border("", text = "The NHS Pharmacy First - Clinical Pathway consultation advanced service was introduced from 31 January 2024. Figures for items issued by through the Clinical Pathway service have been included in the overall counts for dispensed items and drug costs for pharmacies in this publication. This also incudes approximately 62,000 items where a consultation has taken place but no drug was issued.", width = "100%")`
::: 
:::: 

### 2.1 Essential services for contractors {.toc-ignore}

#### `r paste0("The number of community pharmacies and appliance contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 1: The number of active community pharmacies and appliance contractors reached a high in 2023/24

```{r total_contractors_chart, out.width="100%"}
figure_1

get_download_button(title = "Download chart data", data = table_figure_1, filename = "figure 1")

```
##### Table

###### Table 1: The number of active community pharmacies and appliance contractors reached a high in 2023/24

```{r total_contractors_table}
knitr::kable(table_figure_1, align = "llr")

get_download_button(title = "Download chart data", data = table_figure_1, filename = "figure 1")
```
</div>

Source: [Summary tables - General Pharmaceutical Services - Table 1](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 12,009 active community pharmacies in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 112 active appliance contractors in 2023/24.</b>", width = "100%")`
:::
::::

In 2023/24, there were 12,009 active community pharmacies, 595 more than in 2022/23 and the highest since 2015/16. 

The number of active appliance contractors in 2023/24 in England was 112, an increase by one compared to 2022/23.

An active community pharmacy or appliance contractor is one that has submitted prescriptions to the NHSBSA for processing in the specified time period. If a pharmacy has opened, submitted a prescription to the NHSBSA and then closed again in the same year, it would still be classed as an active pharmacy.

When a pharmacy contract changes providers it can remain in the same premises but may be given a new organisation code. This measure uses the pharmacy organisation code to determine active pharmacies. In 2023/24 an increased number of pharmacy contracts changed providers compared to previous years. This may inflate the number of active pharmacies in the year compared to the number of pharmacy premises.

Monthly data relating to Dispensing Contractors can be accessed via the [NHSBSA Open Data Portal.](https://opendata.nhsbsa.net/theme/dispensing-contractors)


#### `r paste0("The total items dispensed by community pharmacies and appliance contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 2: The number of items dispensed by community pharmacies and appliance contractors increased in 2023/24

```{r items_dispensed, out.width="100%"}
figure_2

get_download_button(title = "Download chart data", data = table_figure_2, filename = "figure 2")
```
##### Table

###### Table 2: The number of items dispensed by community pharmacies and appliance contractors increased in 2023/24

```{r items_dispensed_table}
knitr::kable(table_figure_2, align = "llr")

get_download_button(title = "Download chart data", data = table_figure_2, filename = "figure 2")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 3 and 10](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b> There were 1.1 billion items dispensed by community pharmacies in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>There were 12 million items dispensed by appliance contractors in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b> Total items for community pharmacies have trended upwards for the last three years.</b>", width = "100%")`
:::
::::

The number of items dispensed by community pharmacies in England between 2022/23 and 2023/24 increased by 3.15% from 1.08 billion to 1.11 billion. Overall, the number of items dispensed is 11.8% higher than the 995 million items dispensed in 2015/16.

The number of items dispensed by appliance contractors in 2023/24 increased by 4.89% from 2022/23. This has increased each year since 2015/16, from 7.91 million to 11.7 million items. This is an increase of 47.5%.

#### `r paste0("The average number of items dispensed per contractor between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 3: The average number of items dispensed decreased for community pharmacies and increased for appliance contractors in 2023/24

```{r avg_items_dispensed, out.width="100%"}
figure_3

get_download_button(title = "Download chart data", data = table_figure_3, filename = "figure 3")
```
##### Table

###### Table 3: The average number of items dispensed decreased for community pharmacies and increased for appliance contractors in 2023/24

```{r avg_items_dispensed_table}
knitr::kable(table_figure_3, align = "llr")

get_download_button(title = "Download chart data", data = table_figure_3, filename = "figure 3")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 3 and 10](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> On average, there were 93,000 items dispensed per contractor by community pharmacies in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>On average, there were 100,000 items dispensed per contractor by appliance contractors in 2023/24.</b>", width = "100%")`
:::
::::

The yearly average number of items dispensed per contractor has been calculated by dividing the total number of items dispensed by the total number of active contractors. This average is a mean. For important information about interpreting this figure please see [section 3](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gphs_annual_2023_24_v001.html#about-these-statistics)

The average number of items dispensed per community pharmacy has decreased by 1.97% from 94,500 in 2022/23 to 92,700 in 2023/24.  This is an increase of 11.3% from the average of 83,300 in 2015/16.

The average yearly items dispensed for appliance contractors has increased each year since 2015/16 and was higher than the average items dispensed for community pharmacies since 2020/21. The average is 104,000 in 2023/24, a 52.8% increase from 2015/16.

#### `r paste0("The number of items dispensed via the Electronic Prescription Service (EPS) for community pharmacies and appliance contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 4: The percentage of items dispensed via EPS increased for community pharmacies and appliance contractors in 2023/24

```{r eps_items_dispensed, out.width="100%"}
figure_4

get_download_button(title = "Download chart data", data = table_figure_4, filename = "figure 4")
```
##### Table

###### Table 4: The percentage of items dispensed via EPS increased for community pharmacies and appliance contractors in 2023/24

```{r eps_items_dispensed_table}
knitr::kable(table_figure_4, align = "lrrr")

get_download_button(title = "Download chart data", data = table_figure_4, filename = "figure 4")
```
</div>
`r get_download_button(title = "Download data table", data = figure_4_data, filename = "figure_4")`
Source: [Summary tables - General Pharmaceutical Services - Tables 3 and 10](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 1.08 billion prescription items dispensed via EPS in  2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>99.7% of contractors dispensed at least one item via EPS in  2023/24.</b>", width = "100%")`
:::
::::

An NHS prescription may take the form of either a paper form presented to a contractor by a patient, or an electronic message sent via the EPS.

In 2023/24, 1.08 billion prescription items were submitted for dispensed by community pharmacies and appliance contractors via EPS. This was 96.1% of all items dispensed in the year. This was an increase of 60.7 percentage points from 2015/16, when 355 million items were dispensed via EPS, 35.4% of all items.

Additionally in 2023/24, 99.7% of all contractors dispensed at least one prescription item submitted via EPS, an increase from 97.5% in 2015/16.


#### `r paste0("The total costs and value of fees received for contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 5: The total costs and value of fees received for contractors increased in 2023/24

```{r costs_and_fees, out.width="100%"}
figure_5

get_download_button(title = "Download chart data", data = table_figure_5, filename = "figure 5")
```
##### Table

###### Table 5: The total costs and value of fees received for contractors increased in 2023/24

```{r costs_and_fees_table}
knitr::kable(table_figure_5, align = "lrr")

get_download_button(title = "Download chart data", data = table_figure_5, filename = "figure 5")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 3, 5 and 10](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The total for drugs and appliance reimbursement costs paid to contractors was £10.2 billion in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The value of single activity fees paid to contractors was £1.4 billion in 2023/24.</b>", width = "100%")`
:::
::::

Figure 5 shows the total value of costs for drugs and appliances reimbursed to contractors, along with the value of all single activity fees (SAF) that have been claimed. This is not the total amount paid to contractors for dispensing an item as many fees are not included in these figures, such as additional fees that apply to controlled drugs. 

Costs reimbursed to contractors increased in 2023/24, for the fifth consecutive year. Costs increased by 4.97% between 2022/23 and 2023/24 from £9.72 billion to £10.2 billion, the highest costs in 9 years.

The value of SAFs received by contractors increased in 2023/24 by 3.04% to £1.44 billion from £1.40 billion in 2022/23. 

It should be noted, prior to December 2016 the figures reported are for professional fees only, and do not include the three additional payments that the SAF consists of.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Single Activity Fee", text = "Contractors can receive many payments for fulfilling a prescription. These include being reimbursed for the basic price of the drug or appliance they have dispensed and receiving fees that relate to the dispensing of that item. There are many fees that can be claimed by a contractor depending on the item that is dispensed. Details of what fees are payable to community pharmacies and appliance contractors can be found in the [Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff).

The SAF is a fixed fee that applies to every prescription item that is dispensed. The SAF was introduced in December 2016 by the Department of Health and Social Care (DHSC) to consolidate a range of payments into one single fee. These payments were:

* the professional fee (also known as dispensing fee)
* practice payment
* repeat dispensing payment
* EPS monthly allowance

The value of the SAF is set by DHSC and can fluctuate throughout the year. It is based upon the forecast of the total number of items dispensed to ensure that the fee delivery remains within the agreed funding envelope for the year. Table 1 displays the value of the SAF and the dates that it has changed.", width = "100%")`
:::
::::

#### **`r paste0("Table ",1,": Dispensing fees, their value, and when they came into effect")`**
``` {r saf_values, out.width="100%"}

table_1_saf
get_download_button(title = "Download data table", data = table_1_data, filename = "table_1")
```

Source: [The Drug Tariff for England and Wales](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff)

### 2.2 Advanced Services for Contractors {.toc-ignore} 


::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12} 
`r infoBox_border("Seasonal influenza vaccination advanced service", text = "In 2015 community pharmacies began providing seasonal influenza vaccinations under a nationally commissioned service by NHS England & Improvement, now NHS England. Each year from September through to March pharmacy contractors can administer flu vaccines to patients and submit a claim to NHSBSA for payment. This includes reimbursement of the cost of the vaccine, plus a fee for providing the service to NHS patients.

The figures in these statistics for flu vaccinations only relate to those administered by community pharmacies, and do not include vaccines issued to GP patients, healthcare workers, or children of primary school age. Official statistics for these are produced by the UK Health Security Agency (UKHSA) and [can be found on their Immunisation webpage](https://www.gov.uk/health-and-social-care/health-protection-immunisation).", width = "100%")`
:::
::::

#### `r paste0("The number of flu vaccinations provided by contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 6: The number of flu vaccinations provided by contractors decreased in 2023/24

```{r flu_items, out.width="100%"}
figure_6

get_download_button(title = "Download chart data", data = table_figure_6, filename = "figure 6")
```
##### Table

###### Table 6: The number of flu vaccinations provided by contractors decreased in 2023/24

```{r flu_items_table}
knitr::kable(table_figure_6, align = "lrr")

get_download_button(title = "Download chart data", data = table_figure_6, filename = "figure 6")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 3.8 million flu vaccines administered by community pharmacies  in  2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 9,200 community pharmacies administering flu vaccines in 2023/24.</b>", width = "100%")`
:::
::::

The number of vaccines administered by pharmacies as part of the Flu Vaccination advanced service decreased in 2023/23 after increasing every year since the service began in 2015/16. In 2023/24 there were 3.77 million vaccines administered by 9,170 community pharmacies, at an average of 412 vaccines per pharmacy. This was a decrease of 24.7% on the 5.01 million vaccines administered in 2023/24.




::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Average number of vaccines", text = "The average number of flu vaccines administered per community pharmacy was calculated by dividing the total number of flu vaccines administered for each year by the total number of community pharmacies that have provided the service. This average is a mean.", width = "100%")`
:::
::::

The average number of flu vaccines administered per community pharmacy that provides the seasonal influenza advanced service follows a similar trend to the total number of flu vaccines administered. In 2023/24 the average number of flu vaccines administered per community pharmacy was 412. This is approximately 5 times the average in 2015/16, which was 83.

#### `r paste0("The cost and fees for flu vaccinations provided by contractors between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure  7: The cost and fees for flu vaccinations decreased in 2023/24

```{r flu_costs_and_fees, out.width="100%"}
figure_7

get_download_button(title = "Download chart data", data = table_figure_7, filename = "figure 7")
```
##### Table

###### Table 7: The cost and fees for flu vaccinations decreased in 2023/24

```{r flu_costs_and_fees_table}
knitr::kable(table_figure_7, align = "lrr")

get_download_button(title = "Download chart data", data = table_figure_7, filename = "figure 7")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The cost of flu vaccines was £49 million in  2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The value of flu vaccine fees was £36 million in 2023/24.</b>", width = "100%")`
:::
::::

In line with the increase in the number of flu vaccines administered each year, the costs reimbursed and fees paid to community pharmacies as part of the advanced service have increased dramatically. The cost of vaccines administered was 13 times higher in 2023/24 than 2015/16, increasing from £3.75 million to £48.5 million.

The value of fees paid to pharmacies for delivering the flu vaccination service was almost 7 times higher in 2023/24 than 2015/16, increasing from £5.44 million to £36.10
million.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("New Medicine Service", text = "The New Medicine Service (NMS) is an Advanced Service within the NHS Community Pharmacy Contractual Framework. The service provides support for people with long-term conditions when they are newly prescribed a medicine, with the goal of improving medicines adherence. It is focused on particular patient groups and conditions. Details of the currently covered patient groups can be found in the [NHS England Service Specification](https://www.england.nhs.uk/publication/advanced-service-specification-nhs-new-medicine-service-nms/).


The fee for NMS varies between £20 and £28 depending on the volume of prescriptions dispensed by a contractor and the number of NMSs undertaken.", width = "100%")`
:::
::::

#### `r paste0("The total New Medicine Services undertaken by community pharmacies between  2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure  8: The total New Medicine Services undertaken by community pharmacies increased in 2023/24

```{r advanced services nms, out.width="100%"}
figure_8

get_download_button(title = "Download chart data", data = table_figure_8, filename = "figure 8")
```
##### Table

###### Table 8: The total New Medicine Services undertaken by community pharmacies increased in 2023/24

```{r advanced services nms_table}
knitr::kable(table_figure_8, align = "lr")

get_download_button(title = "Download chart data", data = table_figure_8, filename = "figure 8")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 4.0 million NMSs claimed in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The number of NMSs claimed in 2023/24 has increased by 42% from 2022/23.</b>", width = "100%")`
:::
::::

NMSs have shown sizable increases for the last three financial years. Thirteen additional conditions were added to the specification list in September 2021.

<!-- During 2021/2022 a catch-up NMS was introduced for patients who were prescribed a new medicine but did not receive an NMS or any additional support during the COVID-19 pandemic. This catch-up service was offered to eligible patients between 1st September 2021 and 31st March 2022. -->

#### `r paste0("The total Cost of New Medicine Services undertaken by community pharmacies between  2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure  9: The total Cost of New Medicine Services undertaken by community pharmacies increased in 2023/24

```{r advanced services nms cost, out.width="100%"}
figure_9

get_download_button(title = "Download chart data", data = table_figure_9, filename = "figure 9")
```
##### Table

###### Table 9: The total Cost of New Medicine Services undertaken by community pharmacies increased in 2023/24

```{r advanced services nms_cost_table}
knitr::kable(table_figure_9, align = "lr")

get_download_button(title = "Download chart data", data = table_figure_9, filename = "figure 9")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The cost of NMSs was £110 million in  2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The cost of NMSs  claimed in 2023/24 has increased by 43% from 2022/23.</b>", width = "100%")`
:::
::::

In line with the increase in the number of NMSs provided, fees paid to community pharmacies as part of the NMS have increased by 43.3% since 2022/23.



::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Appliance Use Review", text = "Appliance Use Review (AUR) is an advanced service in the NHS Community Pharmacy Contractual Framework. AURs can be carried out by a pharmacist or a specialist nurse in the pharmacy or at the patient’s home.

AURs should improve the patient’s knowledge and use of any ‘specified appliance‘ by:

* establishing the way the patient uses the appliance and the patient’s experience of such use
* identifying, discussing and assisting in the resolution of poor or ineffective use of the appliance by the patient
* advising the patient on the safe and appropriate storage of the appliance
* advising the patient on the safe and proper disposal of the appliances that are used or unwanted

A fee of £28 is payable for an AUR conducted on the premises of a contractor, or by telephone or video link where clinically appropriate<sup>1</sup>, managed by the pharmacy or appliance contractor. A fee of £54 is payable for a review conducted at the patient’s home. When carrying out AURs for multiple patients living at the same property in the same 24 hours, only the first AUR is paid at the home rate. Subsequent AURs paid at the lower premises rate.

The total number of AURs that an appliance or pharmacy contractor may claim fees for is limited to one for every 35 Part IXA (qualifying items), Part IXB and Part IXC prescription items dispensed in the financial year of the claim.

<small>1. The telephone or video link option was introduced as of 1 September 2020.</small>", width = "100%")`
:::
::::

#### `r paste0("The total Appliance Use Reviews (AUR) provided between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 10: The total Appliance Use Reviews (AUR) increased in 2023/24

```{r advanced services aur, out.width="100%"}
figure_10

get_download_button(title = "Download chart data", data = table_figure_10, filename = "figure 10")
```
##### Table

###### Table 10: The total Appliance Use Reviews (AUR) increased in 2023/24

```{r advanced services aur_table}
knitr::kable(table_figure_10, align = "lrrr")

get_download_button(title = "Download chart data", data = table_figure_10, filename = "figure 10")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 7 and 11](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 101,000 AURs provided in 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> In 2023/24 the number of AURs increased by 35% compared to 2022/23.</b>", width = "100%")`
:::
::::

There were 101,000 AURs provided in 2023/24, an increase of 35.2% on the 74,700 provided in 2022/23.

Both community pharmacies and appliance contractors may offer this advanced service. A breakdown by contractor type and AUR type can be found in Tables 7 and 11 in the accompanying summary tables.

#### `r paste0("The total Cost of Appliance Use Reviews (AUR) provided between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 11: The total Cost of Appliance Use Reviews (AUR) increased in 2023/24

```{r advanced services aur cost, out.width="100%"}
figure_11

get_download_button(title = "Download chart data", data = table_figure_11, filename = "figure 11")
```
##### Table

###### Table 11: The total Cost of Appliance Use Reviews (AUR) increased in 2023/24

```{r advanced services aur_costtable}
knitr::kable(table_figure_11, align = "lrrr")

get_download_button(title = "Download chart data", data = table_figure_11, filename = "figure 11")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 7 and 11](`r config$fy_excel_link`)

<!-- ::::{.row style="display: flex; padding-bottom: 15px;"} -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of AURs provided in 2023/24 was £4.6 million.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b>The cost of AURs provided in 2023/24 increased by 39% compared to 2022/23.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::: -->
The cost of AURs provided in 2023/24 was £4.59 million, an increase of 38.8% compared to the cost of £3.31 million in 2022/23.


::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Stoma Appliance Customisation", text = "Stoma Appliance Customisation (SAC) is an advanced service in the NHS Community Pharmacy Contractual Framework. The service involves the customisation of stoma appliances, based on the patient’s measurements or a template. The aim of the service is to ensure proper use and comfortable fitting of the stoma appliance and to improve the duration of usage, thereby reducing waste. The stoma appliances that can be customised are listed in Part IXC of the Drug Tariff.

A fee of £4.32 is payable for every Part IXC prescription item that can be customised as denoted in the Drug Tariff. Pharmacy and appliance contractors do not have to claim the fee.  The fee will automatically be paid to pharmacy and appliance contractors in England who have registered their intent to provide stoma appliance customisation services to the NHSBSA.", width = "100%")`
:::
::::

#### `r paste0("The total Stoma Appliance Customisation (SAC) provided between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 12: The total Stoma Appliance Customisation (SAC) increased in 2023/24

```{r advanced services sac, out.width="100%"}
figure_12

get_download_button(title = "Download chart data", data = table_figure_12, filename = "figure 12")
```
##### Table

###### Table 12: The total Stoma Appliance Customisation (SAC) increased in 2023/24

```{r advanced services sac_table}
knitr::kable(table_figure_12, align = "llr")

get_download_button(title = "Download chart data", data = table_figure_12, filename = "figure 12")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 7 and 11](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> There were 1.8 million SACs provided by appliance contractors in  2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 3,000 SACs provided by community pharmacies in  2023/24. </b>", width = "100%")`
:::
::::

The total number of SACs has increased by 5.06% from 1.69 million to 1.78 million between 2022/23 and 2023/24. 

#### `r paste0("The total Cost of Stoma Appliance Customisation (SAC) provided between 2015/16 and 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 13: The total Cost of Stoma Appliance Customisation (SAC) increased in 2023/24

```{r advanced services sac cost, out.width="100%"}
figure_13

get_download_button(title = "Download chart data", data = table_figure_13, filename = "figure 13")
```
##### Table

###### Table 13: The total Cost of Stoma Appliance Customisation (SAC) increased in 2023/24

```{r advanced services sacCOST_table}
knitr::kable(table_figure_13, align = "llr")

get_download_button(title = "Download chart data", data = table_figure_13, filename = "figure 13")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Tables 7 and 11](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The cost of SACs provided in  2023/24 was £7.7 million.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The cost of SACs has increased by 5.1% between 2022/23 and 2023/24.</b>", width = "100%")`
:::
::::

As the SAC is a set payment that has not changed since its introduction, the percentage increase for volume and cost is the same.


::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Community Pharmacist Consultation Services", text = "The NHS Community Pharmacist Consultation Service (CPCS) Advanced Service began on 29 October 2019 to allow patients to have a same day community pharmacist appointment for a minor illness or an urgent supply of a regular medicine.

The service is intended to improve access to services and provide  treatment closer to patients’ homes by harnessing the skills and medicines knowledge of pharmacists, and relieve pressure on GP appointments and emergency departments. 

From 1 November 2020, general practices have been able to refer patients for a minor illness consultation via CPCS, once a local referral pathway has been agreed. 

From April 2023, referrals from other urgent and emergency (UEC) settings are now included as part of the NHS CPCS.

The fee for providing the CPCS is £14 (including VAT) per consultation.

No data prior to 2019/20 is available for Community Pharmacy Consultation Services.

The CPCS is being replaced with the NHS Referrals for Minor Illness & Urgent Medicines Supply element of the Pharmacy First Service (PFS) which commenced in January 2024.", width = "100%")`
:::
::::

#### `r paste0("The total Cost of Community Pharmacy Consultation Services 2019/20 to 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 14: The total Cost of Community Pharmacy Consultation Services (CPCS) increased in 2023/24

```{r advanced services cpcs, out.width="100%"}
figure_14

get_download_button(title = "Download chart data", data = table_figure_14, filename = "figure 14")
```
##### Table

###### Table 14: The total Cost of Community Pharmacy Consultation Services (CPCS) increased in 2023/24

```{r advanced services cpcs_table}
knitr::kable(table_figure_14, align = "lr")

get_download_button(title = "Download chart data", data = table_figure_14, filename = "figure 14")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 8](`r config$fy_excel_link`)

<!-- ::::{.row style="display: flex; padding-bottom: 15px;"} -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of CSPCs provided in  2023/24 was £26 million.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of CSPCs has increased by 48% between 2022/23 and 2023/24.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::: -->
The cost of CPCSs provided in  2023/24 was £26.3 million, a 48% increased on the cost of £17.8 million in 2022/23.

#### `r paste0("The total Cost of Drugs provided during Community Pharmacy Consultation Services 2019/20 to 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 15: The total Cost of Drugs provided during Community Pharmacy Consultation Services (CPCS) increased in 2023/24

```{r advanced services cpcscost, out.width="100%"}
figure_15

get_download_button(title = "Download chart data", data = table_figure_15, filename = "figure 15")
```
##### Table

###### Table 15: The total Cost of Drugs provided during Community Pharmacy Consultation Services (CPCS) increased in 2023/24

```{r advanced services cpcscost_table}
knitr::kable(table_figure_15, align = "lr")

get_download_button(title = "Download chart data", data = table_figure_15, filename = "figure 15")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 8](`r config$fy_excel_link`)

<!-- ::::{.row style="display: flex; padding-bottom: 15px;"} -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of drugs provided for CSPCs provided in  2023/24 was £5.3 million.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of drugs provided for CSPCs has increased by 71% between 2022/23 and 2023/24.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::: -->
The cost of drugs provided for CPCSs provided in  2023/24 was £5.34 million, a 71.4% increase on the £3.12 million in 2022/23.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Smoking Cessation Services", text = "The Smoking Cessation service (SCS) is an advanced service introduced in March 2022. Patients are discharged with their consent to the community pharmacy of their choice to continue smoking cessation treatment began under an NHS Trust.

The SCS is limited to 12 weeks of treatment including any undertaken at the NHS Trust. 

The fees paid for providing the SCS are £30 for the first consultation, £10 for each interim consultation and £40 for the final consultation. 

Consultations will usually be every two weeks in line with the supply of Nicotine Replacement Therapy (NRT) which can be for a maximum of a 2-week period. 

No data prior to March 2022 is available for Smoking Cessation Services.

Further information including which NRT products can be supplied as part of the service can be found [on the NHSBSA website](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/dispensing-contractors-information/nhs-smoking-cessation-service-referral-secondary-care-community-pharmacy).", width = "100%")`
:::
::::

#### `r paste0("The total Cost of Consultations and Drugs provided during Smoking Cessation Services 2022/23 to 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 16: The total Cost of Consultations and Drugs provided during Smoking Cessation Services (SCS) increased in 2023/24

```{r advanced services scs, out.width="100%"}
figure_16

get_download_button(title = "Download chart data", data = table_figure_16, filename = "figure 16")
```
##### Table

###### Table 16: The total Cost of Consultations and Drugs provided during Smoking Cessation Services (SCS) increased in 2023/24

```{r advanced services scs_table}
knitr::kable(table_figure_16, align = "lrr")

get_download_button(title = "Download chart data", data = table_figure_16, filename = "figure 16")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

<!-- ::::{.row style="display: flex; padding-bottom: 15px;"} -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of consultations for SCS provided in  2023/24 was £130,000.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-6} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of NRT products provided for SCS in 2023/24 was £180,000.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::: -->

The cost of consultations for SCS provided in  2023/24 was £127,000 with the cost of NRT products provided for SCS in 2023/24 being  £180,000. These are both a seven fold increase on the 2022/23 figures when the SCS was introduced.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Pharmacy Contraceptive Services", text = "The Pharmacy Contraceptive Services (PCS) is an advanced service that commenced in April 2023, allowing the ongoing supply of oral contraception (OC) from community pharmacies.

From December 2023, the service included both initiation and ongoing supply of OC. 

The fee for providing the PCS is £18 (including VAT) per consultation.

No data prior to April 2023 is available for Pharmacy Contraceptive Services.", width = "100%")`
:::
::::

#### `r paste0("The total Cost of Consultations and Drugs provided during Pharmacy Contraceptive Services 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 17:The total Cost of Consultations and Drugs provided during Pharmacy Contraceptive Services in 2023/24

```{r advanced services pcs, out.width="100%"}
figure_17

get_download_button(title = "Download chart data", data = table_figure_17, filename = "figure 17")
```
##### Table

###### Table 17: The total Cost of Consultations and Drugs provided during Pharmacy Contraceptive Services in 2023/24

```{r advanced services pcs_table}
knitr::kable(table_figure_17, align = "lrr")

get_download_button(title = "Download chart data", data = table_figure_17, filename = "figure 17")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 6](`r config$fy_excel_link`)

<!-- ::::{.row style="display: flex; padding-bottom: 15px;"} -->
<!-- :::{.col-md-4} -->
<!-- `r infoBox_no_border("", text = "<b> The cost of consultations for PCS provided in  2023/24 was £374,000.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-4} -->
<!-- `r infoBox_no_border("", text = "<b>The cost of OC producsts provided for PCS in 2023/24 was £1.3 million.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::{.col-md-4} -->
<!-- `r infoBox_no_border("", text = "<b>Uptake has trended upwards across the year and the spike in December 2023 is in line with the PCS expansion to include intiation as well as ongoing contraception.</b>", width = "100%")` -->
<!-- ::: -->
<!-- :::: -->
The cost of consultations for PCS provided in 2023/24 was £1.25 million and the cost of OC products provided for PCS in 2023/24 was £374,000.

Figure 17 shows an upwards trend in uptake after November 2023 in line with the PCS expansion to include initiation as well as ongoing contraception. However in the same time period an error was made in December and January payments. These payments were later reclaimed by the NHSBSA. This has not been reflected in this data and therefore the increase in uptake in December and January was not as significant as shown here.

### 2.3 Enhanced Services for Contractors {.toc-ignore} 

Data is not available for most Local Enhanced Services because the NHSBSA holds data of amounts paid but not which locally commissioned service these payments were assigned to. An exception is the COVID-19 vaccination enhanced service which began as a local enhanced service.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("COVID-19  vaccination enhanced service", text = "Towards the end of 2020/21 community pharmacies began providing COVID-19 vaccinations as a local enhanced service to deliver Phase 1 and Phase 2 cohorts  of the Joint Committee on Vaccination and Immunisation (JCVI). A further enhanced service was commissioned to deliver Phase 3, which was provided alongside the seasonal influenza vaccination programme which runs each autumn.

The figures in these statistics for COVID-19 vaccinations only relate to those administered by community pharmacies. A dashboard including COVID-19 vaccination data is available from the UKHSA [on their website](https://coronavirus.data.gov.uk/).", width = "100%")`
:::
::::

#### `r paste0("The total Costs for COVID-19 vaccinations fees provided by contractors in 2020/21 to 2023/24")` {.tabset}

<div class = "tabset">

##### Chart

###### Figure 18: The total Costs for COVID-19 vaccinations fees provided by contractors decreased in 2023/24

```{r covidvaccine_costs_and_fees, out.width="100%"}
figure_18

get_download_button(title = "Download chart data", data = table_figure_18, filename = "figure 18")
```
##### Table

###### Table 18: The total Costs for COVID-19 vaccinations fees provided by contractors decreased in 2023/24

```{r covidvaccine_costs_and_fees_table}
knitr::kable(table_figure_18, align = "lr")

get_download_button(title = "Download chart data", data = table_figure_18, filename = "figure 18")
```
</div>
Source: [Summary tables - General Pharmaceutical Services - Table 9](`r config$fy_excel_link`)

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The costs of COVID-19 vaccines fees in 2023/24 were £75 million.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> The costs of COVID-19 vaccines fees have decreased by 30% between 2022/23 and 2023/24.</b>", width = "100%")`
:::
::::

The COVID-19 vaccines made available under this enhanced service are provided to the pharmacy contractor free of charge. Therefore no data on the cost of the vaccine is held by the NHSBSA.

---

## 3. About these statistics

The General Pharmaceutical Services publication (GPhS) is an official accredited statistic release. Official accredited statistic status means this release complies with the [Code of Practice for Statistics](https://code.statisticsauthority.gov.uk/), showing trustworthiness, quality, and value. 

These official statistics were independently reviewed by the (OSR) in December 2021. You can read more about this confirmation and the assessment of these statistics [on the OSR website](https://osr.statisticsauthority.gov.uk/correspondence/ed-humpherson-to-matthew-wilson-confirmation-of-national-statistics-designation-for-statistics-on-general-pharmaceutical-services-in-england/).


### 3.1. Changes made to this publication {.toc-ignore}

#### Changes to the statistical narrative 

The advanced services Smoking Cessation Service (SCS) and Pharmacy Contraceptive Service (PCS) have been added to the narrative.

#### Additions to the summary tables

Data regarding the advanced services Pharmacy Contraceptive Service (PCS) and Pharmacy First Service (PFS) have been added to the summary tables.


### 3.2. Accuracy of these statistics {.toc-ignore}

Further information on the accuracy of the data used in this publication is available in the [Background Information and Methodology supporting document](`r config$background_link`). Details regarding the impact of post payment verification (PPV) and other checks carried out by NHS Prescription Services (a division of NHSBSA) on the statistics are available in the [Impacts of errors identified on the accuracy of the General Pharmaceutical Services](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/gphs/gps-2021-accuracy-v001.html) supporting document that accompanied the 2020/21 publication.


### 3.3. Uses of these statistics {.toc-ignore}

These statistics can be used to understand the activity of community pharmacies and appliance contractors in England, including essential and advanced services provided to NHS patients. They cannot be used to provide the total number or cost of prescription items dispensed in England in the community as they do not contain dispensing from all sources. This can be obtained from another NHSBSA official accredited statistic publication, [Prescription Cost Analysis](https://www.nhsbsa.nhs.uk/statistical-collections/prescription-cost-analysis-england).


### 3.4. Geographies included in this publication {.toc-ignore}

The geographies used in this publication are based upon NHSBSA administrative records, not geographical health boundaries as defined by the Office for National Statistics (ONS). These administrative records more closely reflect the operational organisation of dispensing contractors than other geographical data sources such as the National Statistics Postcode Lookup (NSPL).

### 3.5 Active pharmacies {.toc-ignore}

An active community pharmacy or appliance contractor is one that has submitted prescriptions to the NHSBSA for processing in the specified time period. If a pharmacy has opened, submitted a prescription to the NHSBSA and then closed again in the same year, it would still be classed as an active pharmacy.

When a pharmacy contract changes providers it can remain in the same premises but may be given a new organisation code. This measure uses the pharmacy organisation code to determine active pharmacies.

Monthly data relating to Dispensing Contractors can be accessed via the [NHSBSA Open Data Portal.](https://opendata.nhsbsa.net/theme/dispensing-contractors)


### 3.6 Planned changes to this publication {.toc-ignore}


NHSBSA plans to expand the scope of this publication in future releases to report on all fees that a contractor can claim for providing NHS services, in order to provide a more complete picture of the total cost of community pharmacy to the public purse. Next year's release will include the below:

The Pharmacy First Service (PFS) is a new advanced service which commenced in January 2024. It will replace the CPCS. The full service will consist of three elements:

* Pharmacy First (clinical pathways), a new element

* Pharmacy First (urgent repeat medicine supply), previously commissioned as the CPCS

* Pharmacy First (NHS referrals for minor illness), previously commissioned as the CPCS

It will be included in this release from the 2024/2025 publication.

The NHS Pharmacy First - Clinical Pathway consultation advanced service was introduced from 31 January 2024. Figures for items issued by through the Clinical Pathway service have been included in the overall counts for dispensed items and drug costs for pharmacies in this publication.  Further work to exclude these items will be undertaken for future iterations of this publication. 

---

## 4. Rounding

The high-level figures in this statistical summary have been rounded where appropriate for clarity, in most cases to three significant figures. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. 

---

## 5. Accessibility

### How to request content in an accessible format {.toc-ignore}

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: accessibility@nhsbsa.nhs.uk

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---


## 6. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics.

You can complete a [short survey about this publication](`r config$stats_survey_link`)  to help us improve the Official Statistics that we produce. All responses will remain anonymous, and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**

NHSBSA - Statistics<br>
NHS Business Services Authority<br>
Stella House<br>
Goldcrest Way<br>
Newburn Riverside<br>
Newcastle upon Tyne<br>
NE15 8NY


**Responsible statistician:** `r config$responsible_statistician`


</main>
